### 2023-07-19 README.Rmd
making compendium

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "Title of your paper goes here"
```

# archmicrobial2023

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/LiYingWang/archmicrobial2023/master?urlpath=rstudio)

This repository contains the data and code for our paper:

> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

## Contents

The **analysis** directory contains:

  - [:file\_folder: paper](/analysis/paper): Quarto source document
    for manuscript. Includes code to reproduce the figures and tables
    generated by the analysis. It also has a rendered version,
    `paper.docx`, suitable for reading (the code is replaced by figures
    and tables in this file)
  - [:file\_folder: data](/analysis/data): Data used in the analysis.
  - [:file\_folder: figures](/analysis/figures): Plots and other
    illustrations
  - [:file\_folder:
    supplementary-materials](/analysis/supplementary-materials):
    Supplementary materials including notes and other documents
    prepared and collected during the analysis.

## How to run in your browser or download and run locally

This research compendium has been developed using the statistical programming
language R. To work with the compendium, you will need
installed on your computer the [R software](https://cloud.r-project.org/)
itself and optionally [RStudio Desktop](https://rstudio.com/products/rstudio/download/).

You can download the compendium as a zip from from this URL:
[master.zip](/archive/master.zip). After unzipping:
- open the `.Rproj` file in RStudio
- run `devtools::install()` to ensure you have the packages this analysis depends on (also listed in the
[DESCRIPTION](/DESCRIPTION) file).
- finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.qmd")` in the R console

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2024-01-29 000-tidy-data.R
edit code

@@ -39,6 +39,9 @@ OTU_V1 <- read.table(here::here("analysis","data","raw_data","V1_asv_table.txt")
 OTU_V6 <- read.table(here::here("analysis","data","raw_data","V6_asv_table.txt"), header = T, sep="\t")
 OTU_ITS1 <- read.table(here::here("analysis","data","raw_data","ITS1_asv_table.txt"), header = T, sep="\t")
 
+check <- OTU_ITS1 %>%
+  select(OTU.ID, taxonomy)
+
 # function to get reads only with ASVs as row names
 reads_only <- function(OTU) {
   OTU_df <- OTU %>%

### 2024-01-29 006-taxa-heatmap.R

@@ -23,7 +23,7 @@ top_freq_genus <- function(df, num){
 target_family_heatmap <- function(df, vec){
 sep_line <- data.frame(x= c(3.5,9.5,15.5), y= c(0.5,0.5,0.5), xend= c(3.5,9.5,15.5), yend= c(Inf,Inf,Inf))
    df %>%
-    filter(Family %in% c("f__Leuconostocaceae", "f__Enterococcaceae","f__Enterobacteriaceae","f__Phycisphaeraceae", "f__Paenibacillaceae")) %>%
+    filter(Family %in% c("f__Enterococcaceae","f__Enterobacteriaceae","f__Phycisphaeraceae", "f__Paenibacillaceae", "f__Leuconostocaceae")) %>%
     group_by(label, Family) %>%
     mutate(sum = sum(count)) %>%
     distinct(label, Family, sum) %>%
@@ -214,86 +214,4 @@ dev.off()
 
 # get the target Family based on references and taxa testing
 V6_target_family_heatmap <- target_family_heatmap(V6_tax_rarefy_df, "V6 region:")
-
-#####################  ITS1: get the top ASVs based on the counts by 4 groups
-ITS1_tax_group_for_heatmap <- top_freq_ASV(ITS1_tax_rarefy_df, 15)
-# plot heat map (ASV level)
-ITS1_heatmap_plot <-
-  ITS1_tax_rarefy_df %>%
-  filter(name %in% c(ITS1_tax_group_for_heatmap)& !Phylum == "unclassified") %>%
-  mutate(Taxa = ifelse(!Species == "unclassified", Species, Genus)) %>%
-  mutate(Taxa = ifelse(Taxa == "unclassified", Family, Taxa)) %>%
-  mutate(Taxa = ifelse(Taxa == "unclassified", Order, Taxa)) %>%
-  mutate(Taxa = ifelse(Taxa == "unclassified", Class, Taxa)) %>%
-  mutate(Taxa = ifelse(Taxa == "unclassified", Phylum, Taxa)) %>%
-  #mutate(Genus = str_remove(Genus, "g__"),
-  # Genus = str_replace(Genus, "(.*)", "*\\1*")) %>% #label = str_replace(label, "([0-9]+)", "\\1\n")
-  mutate(z_scores = (count-mean(count))/sd(count)) %>% #heatmap based on count table (z-score)
-  ggplot(aes(x = label, y = Taxa, fill = z_scores)) +
-  geom_tile() + #color='White', linewidth=0.1
[…]

### 2024-01-29 007-taxa-ITS1-heatmap.R
work on heatmap

@@ -1,0 +1,103 @@
+# run 004 code first to get rarefied data
+#####################  ITS1: get the top ASVs based on the counts by 4 groups
+ITS1_tax_group_for_heatmap <- top_freq_ASV(ITS1_tax_rarefy_df, 15)
+
+# plot heat map (ASV level)
+ITS1_heatmap_plot <-
+  ITS1_tax_rarefy_df %>%
+  filter(name %in% c(ITS1_tax_group_for_heatmap)& !Phylum == "unclassified") %>%
+  mutate(Taxa = ifelse(!Species == "unclassified", Species, Genus)) %>%
+  mutate(Taxa = ifelse(Taxa == "unclassified", Family, Taxa)) %>%
+  mutate(Taxa = ifelse(Taxa == "unclassified", Order, Taxa)) %>%
+  mutate(Taxa = ifelse(Taxa == "unclassified", Class, Taxa)) %>%
+  mutate(Taxa = ifelse(Taxa == "unclassified", Phylum, Taxa)) %>%
+  #mutate(Genus = str_remove(Genus, "g__"),
+  # Genus = str_replace(Genus, "(.*)", "*\\1*")) %>% #label = str_replace(label, "([0-9]+)", "\\1\n")
+  mutate(z_scores = (count-mean(count))/sd(count)) %>% #heatmap based on count table (z-score)
+  ggplot(aes(x = label, y = Taxa, fill = z_scores)) +
+  geom_tile() + #color='White', linewidth=0.1
+  scale_fill_viridis_c(begin= 0.1, end= 0.9) + #direction= -1; #scale_fill_distiller(palette = "RdPu") +
+  theme(axis.text.y = element_text(size = 5)) +
+  coord_equal() +
+  scale_y_discrete(position = "right") +
+  labs(x = NULL, y = NULL, title = "ITS1 region: ASV level") +
+  geom_segment(data=sep_line, aes(x,y,xend=xend, yend=yend), linewidth=0.5, color= "white", inherit.aes=F) +
+  theme_minimal() +
+  theme(axis.text.y = element_markdown(), legend.position="left")
+
+CairoPNG(here::here("analysis","figures","04-ITS1-ASV-heatmap.png"), width = 11, height = 6, dpi = 360, units = "in")
+ITS1_heatmap_plot
+dev.off()
+
+#Genus level and heatmap
+ITS1_tax_Genus_for_heatmap <- top_freq_genus(ITS1_tax_rarefy_df, 5)
[…]

### 2024-01-29 analysis/paper/paper.qmd
work on introduction

---
title: "Characterization of the microbial and fungal communities in ancient pottery from Neolithic southwestern China"
author:
  - Li-Ying Wang:
      correspondence: "yes"
      email: liying15@gmail.com
      orcid: 0000-0001-9060-2125
      institute:
        - AS
  - Kuei-Chen Lin:
      institute: AS
      orcid: 0000-0003-1689-0558
  - Zhiqing Zhou:
      orcid: 0000-0003-1689-0551
      institute:
        - name: Chengdu Institute of Cultural Relics and Archaeology
  - Yin-Ru Chiang:
      correspondence: "yes"
      institute: AS
      orcid: 0000-0003-1689-0559
institute:
  - AS:
      name: Academia sinica
      address: 130, Section 2, Academia Rd, Nangang District, Taipei City, Taiwan
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The actual document text starts here: -->

# Introduction
Ceramics are the most common artifacts unearthed from archaeological sites, offering valuable insights into diverse anthropological facets encompassing cultural, social, and economic dimensions. Foodway is one of the topics that can can explored based on food residues absorbed into the matrices of ceramics. Food residues are useful sources to explore pottery use and past culinary practices. This analytical pursuit heavily relies on contemporary scientific methodologies. For example, organic residues, particularly lipids, extracted from pottery through a series of chemical techniques enable the identification of terrestrial animals, marine resources, and plants [@Evershed2008; @Evershed2002; @Evershed2008marine; @Craig2011; @Heron2016]. Furthermore, the identification of foodstuffs contained in storage vessels can be achieved through ancient DNA analysis, providing direct evidence of pottery content at the genes or species levels [@Foley2012; @Hansson2008; @Robinson2017]. These analytical approaches underscore the significance of preserving the condition of food residues within archaeological contexts, ensuring the efficacy of these scientific methodologies.

Recently, the application of next-generation sequencing in microbiology has brought a novel biomolecular approach to examine ancient human activity through microorganisms present in archaeological samples, including human remains [@Cano2000; @Ubaldi1998; @Ckimsis2023], teeth [@Adler2013; @Warinner2015; @De2013; @Preus2011], archaeological deposits, and minerals [@Margesin2017; @Siles2018; @Lenehan2017; @Boivin2023]. Such microbial archaeology could also be useful for studying food residues adhering to or absorbed into pottery matrix. Based on the assumption that food spoilage mainly results from microbial activity, the microorganisms preserved in pottery within burial contexts could be related to the last prehistoric use of the pot, such as food that was cooked in it. For example, microorganisms including *Proteus spp.*, *Weissella spp.*, *Pseudomonas spp.*, and *Enterobacteriaceae* are frequently associated with meat [@Tassew2010; @Stellato2016], while *Aspergillus spp.*, *Penicillium spp.*, *Fusarium spp.*, *Rhizopus spp.* are commonly found in association with rice [@Li2022; @Bertuzzi2019; @Tahir2012]. Moreover, *Acetobacter*, *Lactobacillus*, *Bacillus*, *Bifidobacterium* are indicative of fermented foods [@Tamang2016; @Xiao2020; @Zhang2018], and *Paenibacillus*, *Chryseobacterium* are largely found in mixed rice-vegetable meals [@Helmond2017; @Nde2021]. Given the potential widespread occurrence of microbial communities across various food categories, more studies are still needed to understand the characteristics of microbial communities in ancient pottery.

In this study, our objective is to investigate the feasibility of using microbial communities as indicators of past food processing in pottery through a comparative analysis of different parts of ceramic samples. We aim to answer the research question is: Can microbial compositions serve as evidence for distinguishing various food categories based on the prevalence of spoilage microorganisms? Our hypothesis is that the microbial composition in absorbed residues within the pottery matrix will exhibit distinctions relevant to food residues when compared to those on the surface of pottery and in surrounding soils. While microorganisms offer indirect evidence of past food practices, our focus is on their potential to provide valuable information that complements existing scientific approaches to analyzing food residues. This innovative perspective sheds light on food practices from a microbial perspective that has not been fully discussed in archaeology.

# Background

# Methods

Microbial DNA extraction for 20 samples (Power soil toolkit)
Focus on bacterial, archaeal and fungal communities
Primer selection for target regions (100-300 bp): 
16S rRNA gene (V1, V4, V6) 
18S rRNA gene (ITS1)
Amplification with PCR
Gel running and collecting the bands of regions
Illumina amplicon sequencing

Amplicons of V1, V4, and V6 of 16S and ITS1 of ITS1.

# Results

Alpha diversity 
Beta diversity
Relative abundance 
Venn diagrams
The abundant taxa at ASV level & the abundant taxa at genus level

Questions:

1. Abundance and relative abundance? Proportion of the general microbial communities across different types of analysis. Richness and Shanon analysis
2. What should I focus on heatmap? those high frequencies? targets? or based on tree? 
3. How to deal with control sets? Subtract? 
4. What is the correction between V1, V4, V6 of 16S according to abundance and taxonomy?
5. Phylogentic-based analysis, phylogenic tree?
6. the distribution of microbiome across different groups of samples?

ANOVA
normal distribution
standard deviation 
Kristall_willis

<!-- Here's some example analysis code: -->

```{r}
#| label: get-data
#| eval: false
# Note the path that we need to use to access our data files when rendering this document
my_data <- read.csv(here::here('analysis/data/raw_data/my_csv_file.csv'))
```

```{r}
#| label: fig-demo-plot
#| fig-cap: "A plot of random numbers"
plot(rnorm(10))
```

@fig-demo-plot shows how we can have a caption and cross-reference for a plot. Note that figure label and cross-references must both be prefixed with `fig-`

```{r}
#| label: demo-inline-code
#| echo: false
x <- round(pi, 2)
```

Here is an example of inline code `r x` in the middle of a sentence.

# Discussion

There are significant differences in the richness and diversity of bacterial and archaeal communities across the four groups: soil > exterior potsherds > interior potsherds > control
Some microbial groups are significantly more abundant in pottery samples, such as Weissella, Escherichia Shigella, Paenibacillus, and Chryseobacterium, and Penicillium
V4 and V6 regions are more ideal for determining the microorganisms in archaeological materials compared with V1 and ITS1 regions

# Conclusion

# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r}
#| label: colophon
#| cache: false

# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```

### 2024-01-30 analysis/paper/paper.qmd
work on cultural background and method

@@
 # Introduction
+
 Ceramics are the most common artifacts unearthed from archaeological sites, offering valuable insights into diverse anthropological facets encompassing cultural, social, and economic dimensions.
-Foodway is one of the topics that can can explored based on food residues absorbed into the matrices of ceramics.
-Food residues are useful sources to explore pottery use and past culinary practices.
+Foodway is one of the topics.
+Food residues absorbed into the matrices of ceramics are useful sources to explore pottery use and past culinary practices.
 This analytical pursuit heavily relies on contemporary scientific methodologies.
@@
 
-In this study, our objective is to investigate the feasibility of using microbial communities as indicators of past food processing in pottery through a comparative analysis of different parts of ceramic samples.
-We aim to answer the research question is: Can microbial compositions serve as evidence for distinguishing various food categories based on the prevalence of spoilage microorganisms?
+In this study, our objective is to investigate the feasibility of using microbial communities as indicators of past food processing in pottery by focusing on the potsherds from the Guijiabao site in southwest China.
+As a long-term occupation site from Neolithic period to historical period, Guijiabao revealed abundant archaeological assemblages and remaines to explore the adaptation of farming communities in a mountain setting.
+The post-excavation analysis of Guijiabao is still in progress.
+This study serves as a pilot study to examine the pottery use from a perspective of microbial archaeology.
+Through a comparative analysis of different parts of ceramic samples and soil samples, we aim to answer the research question: Can microbial compositions serve as evidence for distinguishing various food categories based on the prevalence of spoilage microorganisms?
 Our hypothesis is that the microbial composition in absorbed residues within the pottery matrix will exhibit distinctions relevant to food residues when compared to those on the surface of pottery and in surrounding soils.
@@
 
-# Background
+# Site Description and Sampling
+
+Guijiabao is located in the Yanyuan Basin in the middle Hengduan Mountains of southwest China.
+The chronological framework of Guijiabao includes the Neolithic period (5000-3700 BP), the Bronze Age (3200-2700 BP), and the historical period spanning from the late North Song to early Ming dynasties (1200-500 BP) based on radiocarbon dates [@Hao2022; @Huan2022].
+Given its location along the route of population movements, Guijiabao has been a place for cultural exchanges with neighboring regions, including the northern and southern parts of the Hengduan Mountains, the Sichuan Basin, and western Yunnan since the Neolithic period.
+This culture interactions is evident in the similarities observed in pottery styles and decorations [@Guijiabao2017a].
+Excavation at Guijiabao revealed various pottery forms, including pots, jars, urns, bowls, and basins.
+Predominantly, the pottery types are characterized by grayish high-tempered fine ware and reddish-brown coarse ware featuring cord-marked or basket-impressed patterns [@Hao2022; @Guijiabao2017a].
+
+As the earliest Neolithic site in the Yanyuan Basin, Guijiabao provides valuable data for exploring the lives of people who have lived in a highland environment (approximately 2400 m above sea level).
+Archaeobotanical evidence and phytolith analyses suggested that mixed crop agriculture, specifically rice and millet, was introduced to this region since the occupation of the site [@Huan2022; @Guijiabao2016].
+Cultivated crops, such as broomcorn millet, foxtail millet, and rice were identified as the primary agricultural practices at the site [@Deng2019; @Guijiabao2016].
+Hunting and fishing could be also part of local subsistence activity, according to the finding of stone hunting implements and net sinkers [@Hao2022; @Guijiabao2020a].
+Examining pottery used in daily life becomes critical in shedding light on food practices, offering more evidence for understanding local subsistence via the lens of food preparation.
+
+As a preliminary investigation of the feasibility of using a microbial approach for detecting evidence of cooking and food types, we sampled five potsherds of typical ceramics from Guijiabao, with a pottery sample from Gaoshan, anorther Neolithic site in southwest China, as a comparison.
+Gaoshan is located in southwest Chengdu Plain, an alluvial plain in Sichuan, and dates back to the early Baodun period, the earliest Neolithic culture (ca.
+4700-3700 BP) in the Chengdu Plain [@Gaoshan2017].
+Archaeological evidence from the Gaoshan site suggests a primary subsistence pattern involving rice cultivation and pig husbandry, supplemented by millet cultivation [@Gaoshan2017; @JiangGaoshan2017; @Lee2020; @He2020].
+In addition to the pottery samples, we collected three soil samples from cultural layers across various excavation units at Guijiabao, alongside two soil samples from Gaoshan.
+After sampling, the potsherds were securely stored in sealed boxes, while soil samples were preserved in sterile vials, both of which were subsequently frozen at -80 °C in preparation for DNA analyses.
 
@@
 
-Microbial DNA extraction for 20 samples (Power soil toolkit)
+To minimize potential sources of contamination from the environments, we cleaned all lab benches and pipettors using 99% ethanol and UV light before sample preparation and DNA extraction.
+These procedures were conducted in the biomolecular laboratory facilities at Academia Sinica, Taiwan.
+Because contemporary contamination could be a significant concern when interpreting the data, we used blank controls during sample preparation, DNA extraction, and DNA amplification to monitor extraneous DNA from surroundings, solutions and reagents, and consumables.
+A total of 12 powdered pottery samples (interior and exterior parts), five soil samples from Guijiabao and Gaoshan, and three negative control samples were tested in this study.
+
+## sample preparation
+
+Each pottery sample was divided into two distinct portions: the interior, representing the middle part of the potsherd matrix, and the exterior surface exposed to the environment.
+This division aimed to facilitate a comparative analysis of the microbial communities in residues absorbed into the potsherds versus those on the surface.
+The collection of approximately 1g of the interior potsherd was executed using a sterile drill operated at a low speed to prevent heat generation.
+The powdered sample was initially gathered on a pre-combustion foil and subsequently transferred into pre-sterilized tubes.
+A parallel sampling procedure was conducted for the surface of the potsherds, with the drill being cleaned between each iteration to minimize cross-contamination.
+For the soil samples, around 1g was collected from the tube containing the soil samples using sterile spatulas.
+
+## DNA extraction and amplification
+
+Total DNA was extracted for each sample using the DNeasy Power Soil Pro Kit (QIAGEN). All DNAs were quantified using Qubit Fluorometer.
+
 Focus on bacterial, archaeal and fungal communities

### 2024-01-31 analysis/paper/paper.qmd
work on method

@@
 This study serves as a pilot study to examine the pottery use from a perspective of microbial archaeology.
-Through a comparative analysis of different parts of ceramic samples and soil samples, we aim to answer the research question: Can microbial compositions serve as evidence for distinguishing various food categories based on the prevalence of spoilage microorganisms?
-Our hypothesis is that the microbial composition in absorbed residues within the pottery matrix will exhibit distinctions relevant to food residues when compared to those on the surface of pottery and in surrounding soils.
+We aim to explore the differences in overall microbial communities, including bacterial, archaeal and fungal communities, through a comparative analysis of different parts of ceramic samples and soil samples.
+Our main research question is: Can microbial compositions serve as evidence for distinguishing various food categories based on the prevalence of spoilage microorganisms?
+We hypothesized that the microbial composition in absorbed residues within the pottery matrix will exhibit distinctions relevant to food residues when compared to those on the surface of pottery and in surrounding soils.
 While microorganisms offer indirect evidence of past food practices, our focus is on their potential to provide valuable information that complements existing scientific approaches to analyzing food residues.
-This innovative perspective sheds light on food practices from a microbial perspective that has not been fully discussed in archaeology.
+This innovative perspective could shed light on food practices from a microbial perspective that has not been fully discussed in archaeology.
 
@@
 
-To minimize potential sources of contamination from the environments, we cleaned all lab benches and pipettors using 99% ethanol and UV light before sample preparation and DNA extraction.
-These procedures were conducted in the biomolecular laboratory facilities at Academia Sinica, Taiwan.
-Because contemporary contamination could be a significant concern when interpreting the data, we used blank controls during sample preparation, DNA extraction, and DNA amplification to monitor extraneous DNA from surroundings, solutions and reagents, and consumables.
-A total of 12 powdered pottery samples (interior and exterior parts), five soil samples from Guijiabao and Gaoshan, and three negative control samples were tested in this study.
+To minimize potential sources of contamination from the environments, we cleaned all lab benches, equipment, and pipettors with 3% bleach, followed by 99% ethanol and UV light.
+The sample preparation and DNA extraction were conducted in the biomolecular laboratory facilities at Academia Sinica, Taiwan.
+Because modern contamination could be a significant concern when interpreting the data, we used blank controls during sample preparation, DNA extraction, and DNA amplification to monitor extraneous DNA from surroundings, solutions, reagents, and consumables.
+A total of 12 powdered pottery samples (interior and exterior parts), 5 soil samples from Guijiabao and Gaoshan, and 3 negative control samples were analyzed in this study.
 
-## sample preparation
+## Sample preparation and DNA extracation
 
-Each pottery sample was divided into two distinct portions: the interior, representing the middle part of the potsherd matrix, and the exterior surface exposed to the environment.
+Each pottery sample was divided into two portions: the interior, representing the middle part of the potsherd matrix, and the exterior surface exposed to the environment.
 This division aimed to facilitate a comparative analysis of the microbial communities in residues absorbed into the potsherds versus those on the surface.
-The collection of approximately 1g of the interior potsherd was executed using a sterile drill operated at a low speed to prevent heat generation.
-The powdered sample was initially gathered on a pre-combustion foil and subsequently transferred into pre-sterilized tubes.
-A parallel sampling procedure was conducted for the surface of the potsherds, with the drill being cleaned between each iteration to minimize cross-contamination.
-For the soil samples, around 1g was collected from the tube containing the soil samples using sterile spatulas.
+Approximately 1.5 g of each sample was collected using a sterile drill operated at a low speed to prevent heat generation.
+The powdered sample was initially gathered on pre-combustion foil and immediately transferred to a sterile Eppendorf tube.
+All the tools were cleaned between each iteration to minimize cross-contamination.
+For the soil samples, around 1g was collected using sterile spatulas and transferred to a sterile Eppendorf tube.
 
-## DNA extraction and amplification
+Total DNA from each sample was extracted in a portion of 300 mg using the DNeasy PowerSoil Pro kit (Qiagen, Germany), and the DNA extracts from the same sample were pooled.
+The concentration of DNA was measured using a Qubit 2.0 fluorometer and standardized to 1 ng μL^−1^.
 
-Total DNA was extracted for each sample using the DNeasy Power Soil Pro Kit (QIAGEN). All DNAs were quantified using Qubit Fluorometer.
+## PCR analyses
 
-Focus on bacterial, archaeal and fungal communities
-Primer selection for target regions (100-300 bp): 
-16S rRNA gene (V1, V4, V6) 
-18S rRNA gene (ITS1)
-Amplification with PCR
-Gel running and collecting the bands of regions
-Illumina amplicon sequencing
+Given the considerable fragmentation of DNA in archaeological contexts, we focused on short regions ranging from 100 to 300 bp, including V1, V4, and V6 of 16S rRNA and ITS1 of 18S rRNA.
+For bacterial communities, the V1 region was amplified using the primers 8F (5'-AGAGTTTGATYMTGGCTCAG-3') and 120R (5'-TTACTCACCCGTICGCCRCT-3') [@].
+For archaeal communities, V4 region was amplified using arch519F (5′-CAGCMGCCGCGGTAA-3′) and arch806R (5′-GGACTACVSGGGTATCTAAT-3′), and V6 region was amplified using 926F (5’-AAACTYAAAKGAATTGACGGG-3’) and 1046R (5’-CGACARCCATGCASCACCT-3’) [@Siles2018].
+For fungal communities, ITS1 region of 18S rRNA was amplified using the primers ITS1F (5’-CTTGGTCATTTAGAGGAAGTAA-3’) and ITS2 (5′-GCTGCGTTCTTCATCGATGC-3′) [@Siles2018].
+Barcoded forward adapters were added to the PCR primers to facilitate sample pooling for library preparations.
+The DNA amplification for bacteria was conducted using the Taq DNA polymerase Master Mix RED (Ampliqon) under the following program: initial denaturation at 95 °C for 1 min, followed by 35 cycles of denaturation at 95 °C for 30 s, primer annealing at 50 °C for 30 s, extension at 72 °C for 30 s, and final elongation at 72 °C for 5 min.
+For fungi, the thermal conditions are the same as for bacteria, with the exception of 40 cycles and annealing temperature at 57 °C for 40 s.
 
-Amplicons of V1, V4, and V6 of 16S and ITS1 of ITS1.
+The DNA fragments in PCR products were purified with the GenepHlow gel extraction kit  (Geneaid Biotech Ltd.) after conducting 2 % agarose gel electrophoresis.
+The amplified products were then pooled and used to build a DNA library using the Illumina TruSeq DNA library preparation protocol.
 

### 2024-02-01 002-rarefy-beta-NMDS.R
add 3D NMDS

@@ -63,10 +63,12 @@ V6_pot_perm <- adonis2(rarefy_V6_reads_only ~ group, data = rarefy_V6_meta_join)
 ITS1_pot_perm <- adonis2(rarefy_ITS1_reads_only ~ group, data = rarefy_ITS1_meta_join)
 
 # Non-metric Multidimensional Scaling (NMDS): collapse all species axes into 2 to visualize the differences between samples
+library(vegan3d)
+library(viridis)
 V4_NMDS <- metaMDS(rarefy_V4_reads_only)
 V1_NMDS <- metaMDS(rarefy_V1_reads_only)
 V6_NMDS <- metaMDS(rarefy_V6_reads_only)
-ITS1_NMDS <- metaMDS(rarefy_ITS1_reads_only)
+ITS1_NMDS <- metaMDS(rarefy_ITS1_reads_only) # k value refers to numbers of dimension
 stressplot(ITS1_NMDS)
 plot(ITS1_NMDS)
 
@@ -87,12 +89,12 @@ NMDS_df_all <- rbind(V1_plot_NMDS_df, V4_plot_NMDS_df, V6_plot_NMDS_df, ITS1_plo
 # NMDS plots
 nmds_plot_all <-
   ggplot(NMDS_df_all,
-         aes(x = NMDS1, y = NMDS2, color = group, shape = group)) +
-  geom_point(size = 3, alpha = 0.8) +
-  stat_ellipse(linetype = 2, size = 1) +
-  labs(title = "NMDS") +
-  scale_color_discrete(name = "Sample group",
-                      labels = c("control", "pot-interior", "pot-exterior", "soil")) +
+         aes(x = NMDS1, y = NMDS2, color = group, shape = group)) + #color = group
+  geom_point(size = 2.5, alpha = 0.9) +
+  #stat_ellipse(linetype = 2, size = 1) +
+  #scale_color_discrete(name = "Sample group",
+                     # labels = c("control", "pot-interior", "pot-exterior", "soil")) +
+  scale_colour_viridis_d(option = "magma", begin= 0.1, end= 0.8, direction = -1) +
   scale_shape_discrete(name = "Sample group",
                        labels = c("control", "pot-interior", "pot-exterior", "soil")) +
   theme_minimal() +
@@ -104,7 +106,28 @@ nmds_plot_all <-
[…]

### 2024-02-01 002-rarefy-beta-NMDS.R
edit NMDS plot

@@ -91,20 +91,24 @@ nmds_plot_all <-
   ggplot(NMDS_df_all,
          aes(x = NMDS1, y = NMDS2, color = group, shape = group)) + #color = group
   geom_point(size = 2.5, alpha = 0.9) +
-  #stat_ellipse(linetype = 2, size = 1) +
+  #stat_ellipse(linetype = 2, size = 0.3) +
   #scale_color_discrete(name = "Sample group",
                      # labels = c("control", "pot-interior", "pot-exterior", "soil")) +
-  scale_colour_viridis_d(option = "magma", begin= 0.1, end= 0.8, direction = -1) +
+  scale_colour_viridis_d(option = "magma", begin= 0.2, end= 0.8, direction = -1) + #option = "magma"
   scale_shape_discrete(name = "Sample group",
                        labels = c("control", "pot-interior", "pot-exterior", "soil")) +
-  theme_minimal() +
+  scale_shape_manual(values = c(10, 17, 15, 19)) +
+  geom_vline(xintercept = 0, linetype="dashed", color="gray30", size= 0.5) +
+  geom_hline(yintercept = 0, linetype="dashed", color="gray30", size= 0.5) +
   facet_wrap(~Region, scales = "free") +
+  scale_x_continuous(expand = c(0.1, 0)) +
+  scale_y_continuous(expand = c(0.1, 0)) +
   theme_minimal() +
   theme(plot.background = element_rect(fill = "white", colour = "white"),
-        strip.text = element_text(size= 12))
+        strip.text = element_text(size= 12)) #strip.background = element_blank()
 
 ggsave(here::here("analysis", "figures", "03-NMDS-plot.png"),
-       width = 10, height = 8, units = "in")
+       width = 8, height = 6, units = "in")
 
 # check 3D NMDS
 ITS1_NMDS <- metaMDS(rarefy_ITS1_reads_only, k=3)

### 2024-02-02 002-rarefy-beta.R
edit code

@@ -121,10 +121,10 @@ V6_pcoa_bray_for_plot <- beta_df(V6_bray, V6_meta)
 ITS1_pcoa_bray_for_plot <- beta_df(ITS1_bray, ITS1_meta)
 
 # calculate the proportion of variance in the data explained by the first two PCoA axes
-V4_bray_eig <- round(c(V4_bray$eig[1]/(sum(V4_bray$eig)), V4_bray$eig[2]/(sum(V4_bray$eig))), 4)*100
-V1_bray_eig <- round(c(V1_bray$eig[1]/(sum(V1_bray$eig)), V1_bray$eig[2]/(sum(V1_bray$eig))), 4)*100
-V6_bray_eig <- round(c(V6_bray$eig[1]/(sum(V6_bray$eig)), V6_bray$eig[2]/(sum(V6_bray$eig))), 4)*100
-ITS1_bray_eig <- round(c(ITS1_bray$eig[1]/(sum(ITS1_bray$eig)), ITS1_bray$eig[2]/(sum(ITS1_bray$eig))), 4)*100
+V4_bray_eig <- round(100*V4_bray$eig/sum(V4_bray$eig), 2)
+V1_bray_eig <- round(100*V1_bray$eig/sum(V1_bray$eig), 2)
+V6_bray_eig <- round(100*V6_bray$eig/sum(V6_bray$eig), 2)
+ITS1_bray_eig <- round(100*ITS1_bray$eig/sum(ITS1_bray$eig), 2)
 
 # create a PCoA bray plot
 library(ggrepel)
@@ -152,7 +152,7 @@ ggsave(here::here("analysis", "figures", "PcoA_bray_rare.png"), width = 6.5, hei
 jac_calculate <- function(df){
   #calculate principal coordinates analysis (Bray-Curtis)
   jac <- vegdist(df, dmethod = "jaccard", binary = T)
-  pcoa_jac <- cmdscale(jac, k = 2, eig = T)
+  pcoa_jac <- cmdscale(jac, k = 2, eig = T, add = T) # add to reduce negative values
   return(pcoa_jac)
 }
 
@@ -198,10 +198,15 @@ V6_pcoa_jac_for_plot <- beta_df(V6_jac, V6_meta)
 ITS1_pcoa_jac_for_plot <- beta_df(ITS1_jac, ITS1_meta)
 
[…]

### 2024-02-02 analysis/paper/paper.qmd
work on result

@@
 
+```{r setting}
+#| echo: false
+# Format integers
+options(scipen=999)
+knit_print.integer = function(x, ...) {
+  prettyNum(x, big.mark=",")
+}
+
+registerS3method(
+  "knit_print", "integer", knit_print.integer,
+  envir = asNamespace("knitr")
+)
+
+```
+
 # Introduction
@@
 This study serves as a pilot study to examine the pottery use from a perspective of microbial archaeology.
-We aim to explore the differences in overall microbial communities, including bacterial, archaeal and fungal communities, through a comparative analysis of different parts of ceramic samples and soil samples.
+We aim to explore the differences in overall microbial communities, including bacterial and fungal communities, through a comparative analysis of different parts of ceramic samples and soil samples.
 Our main research question is: Can microbial compositions serve as evidence for distinguishing various food categories based on the prevalence of spoilage microorganisms?
@@
 The sample preparation and DNA extraction were conducted in the biomolecular laboratory facilities at Academia Sinica, Taiwan.
-Because modern contamination could be a significant concern when interpreting the data, we used blank controls during sample preparation, DNA extraction, and DNA amplification to monitor extraneous DNA from surroundings, solutions, reagents, and consumables.
+Because modern contamination could be a significant concern when interpreting the data, we used negative controls during sample preparation, and reagent blanks for DNA extraction, and PCR amplification to monitor extraneous DNA from surroundings, reagents, and consumables.
 A total of 12 powdered pottery samples (interior and exterior parts), 5 soil samples from Guijiabao and Gaoshan, and 3 negative control samples were analyzed in this study.
@@
 
-Given the considerable fragmentation of DNA in archaeological contexts, we focused on short regions ranging from 100 to 300 bp, including V1, V4, and V6 of 16S rRNA and ITS1 of 18S rRNA.
-For bacterial communities, the V1 region was amplified using the primers 8F (5'-AGAGTTTGATYMTGGCTCAG-3') and 120R (5'-TTACTCACCCGTICGCCRCT-3') [@].
-For archaeal communities, V4 region was amplified using arch519F (5′-CAGCMGCCGCGGTAA-3′) and arch806R (5′-GGACTACVSGGGTATCTAAT-3′), and V6 region was amplified using 926F (5’-AAACTYAAAKGAATTGACGGG-3’) and 1046R (5’-CGACARCCATGCASCACCT-3’) [@Siles2018].
-For fungal communities, ITS1 region of 18S rRNA was amplified using the primers ITS1F (5’-CTTGGTCATTTAGAGGAAGTAA-3’) and ITS2 (5′-GCTGCGTTCTTCATCGATGC-3′) [@Siles2018].
+Given the fragmentation of DNA in archaeological contexts, we targeted single hypervariable regions, V1, V4, and V6 of 16S rRNA, and ITS1 of 18S rRNA, to cover both bacterial and fungal communities.
+The V1 region was amplified using the primers 8F (5'-AGAGTTTGATYMTGGCTCAG-3') and 120R (5'-TTACTCACCCGTICGCCRCT-3') [@Adler2013], the V4 region was amplified using arch519F (5′-CAGCMGCCGCGGTAA-3′) and arch806R (5′-GGACTACVSGGGTATCTAAT-3′), and the V6 region was amplified using 926F (5’-AAACTYAAAKGAATTGACGGG-3’) and 1046R (5’-CGACARCCATGCASCACCT-3’) [@Adler2013; @Siles2018].
+For fungal communities, the region ITS1 of 18S rRNA was amplified using the primers ITS1F (5’-CTTGGTCATTTAGAGGAAGTAA-3’) and ITS2 (5′-GCTGCGTTCTTCATCGATGC-3′) [@Siles2018].
 Barcoded forward adapters were added to the PCR primers to facilitate sample pooling for library preparations.
 The DNA amplification for bacteria was conducted using the Taq DNA polymerase Master Mix RED (Ampliqon) under the following program: initial denaturation at 95 °C for 1 min, followed by 35 cycles of denaturation at 95 °C for 30 s, primer annealing at 50 °C for 30 s, extension at 72 °C for 30 s, and final elongation at 72 °C for 5 min.
-For fungi, the thermal conditions are the same as for bacteria, with the exception of 40 cycles and annealing temperature at 57 °C for 40 s.
+For fungi, the thermal conditions are the same as for bacteria, with the exception of 40 cycles and primer annealing temperature at 57 °C for 40 s.
 
@@
 
+## DNA library construction and sequencting
+
+## Taxonomic classification
+
 # Results
+
+```{r get-data}
+#| label: get-data
+#| echo: false
+#| cache: true
+library(tidyverse)
+library(readxl)
+# read original count data
+OTU_V4 <- read.table(here::here("analysis","data","raw_data","V4_asv_table.txt"), header = T, sep="\t")
+OTU_V1 <- read.table(here::here("analysis","data","raw_data","V1_asv_table.txt"), header = T, sep="\t")
+OTU_V6 <- read.table(here::here("analysis","data","raw_data","V6_asv_table.txt"), header = T, sep="\t")
+OTU_ITS1 <- read.table(here::here("analysis","data","raw_data","ITS1_asv_table.txt"), header = T, sep="\t")
+
+OTU_reads_total <- function(df) {
+  sum_read <- df %>%
+    rowwise() %>% 
+    mutate(sum_read = sum(across(where(is.numeric)))) %>% 
+    select(sum_read)
+  return(sum(sum_read))
+}
+
+V4_total_reads <- OTU_reads_total(OTU_V4)
+V1_total_reads <- OTU_reads_total(OTU_V1)
+V6_total_reads <- OTU_reads_total(OTU_V6)
+ITS1_total_reads <- OTU_reads_total(OTU_ITS1)
+total_reads <- V4_total_reads + V1_total_reads + V6_total_reads + ITS1_total_reads
+
+# read rarefied data
+rarefy_V4 <- read.csv(here::here("analysis","data","derived_data", "V4_shared_format.1.subsample.60216.csv"), sep = ",")
+rarefy_V1 <- read.csv(here::here("analysis","data","derived_data","V1_shared_format.1.subsample.13658.csv"), sep = "")
+rarefy_V6 <- read.csv(here::here("analysis","data","derived_data", "V6_shared_format.1.subsample.28402.csv"), sep = ",")
+rarefy_ITS1 <- read.csv(here::here("analysis","data","derived_data", "ITS1_shared_format.1.subsample.387.csv"), sep = ",")
+
+rarefy_reads <- function(df) {
+  sum_read <- df %>%
+    rowwise() %>% 
+    mutate(sum_read = sum(across(starts_with("ASV")))) %>% 
+    select(sum_read)
+  return(sum(sum_read))
+}
+
+V4_rare_reads <- rarefy_reads(rarefy_V4)
+V1_rare_reads <- rarefy_reads(rarefy_V1)
+V6_rare_reads <- rarefy_reads(rarefy_V6)
+ITS1_rare_reads <- rarefy_reads(rarefy_ITS1)
+rare_reads <- V4_rare_reads + V1_rare_reads + V6_rare_reads + ITS1_rare_reads
+
+V4_OTU_num <- unique(rarefy_V4$numOtus)
+V1_OTU_num <- unique(rarefy_V1$numOtus)
+V6_OTU_num <- unique(rarefy_V6$numOtus)
+ITS1_OTU_num <- unique(rarefy_ITS1$numOtus)
+total_OTU <- V4_OTU_num + V1_OTU_num + V6_OTU_num + ITS1_OTU_num
+```
+
+## Alpha diversity analysis of bacterial and fungal communities
+Illumina analysis generated a total of `r total_reads` sequences that were normalized to `r rare_reads` after rarefaction [@Schloss2024].
+Soil samples have the highest shannon diversity index and richness, followed by pottery exterior samples, and pottery interior samples, and negative control.
+
+## Taxonomic composition
+
+## Beta diversity analysis
 
@@
 <!-- Here's some example analysis code: -->
-
-```{r}
-#| label: get-data
-#| eval: false
-# Note the path that we need to use to access our data files when rendering this document
-my_data <- read.csv(here::here('analysis/data/raw_data/my_csv_file.csv'))
-```
 

### 2024-02-03 001-rarefy-alpha.R
add pairwise comparison

@@ -68,6 +68,12 @@ rich_V1_test  <- richness_test(rarefy_V1_reads_only, V1_meta)
 rich_V6_test  <- richness_test(rarefy_V6_reads_only, V6_meta)
 rich_ITS1_test  <- richness_test(rarefy_ITS1_reads_only, ITS1_meta)
 
+pair_richness_test <- function (df1, df2){
+  rich <- specnumber(df1) # the number of species within each sample
+  pair_rich_aov <- pairwise.t.test(rich ~ group, data =df2, p.adj = "none") # analysis of variance between groups
+  return(c(rich, summary(rich), summary(rich_aov)))
+}
+
 sppdiv_aov_V4 <- aov(diversity(rarefy_V4_reads_only) ~ group, data =V4_meta)
 sppdiv_aov_V1 <- aov(diversity(rarefy_V1_reads_only) ~ group, data =V1_meta)
 sppdiv_aov_V6 <- aov(diversity(rarefy_V6_reads_only) ~ group, data =V6_meta)
@@ -80,7 +86,7 @@ alpha_cal_df <- function(df1, df2){
   evenness <- diversity(df1)/log(specnumber(df1))
   shannon <- diversity(df1, index = "shannon")
   alphadiv <- cbind(df2, t(richness), shannon, evenness) #combine all in one
-  rm(richness, evenness, shannon) #remove the unnecessary data/vector
+  #rm(richness, evenness, shannon) #remove the unnecessary data/vector
   return(alphadiv)
 }
 
@@ -88,6 +94,15 @@ V4_alphadiv <- alpha_cal_df(rarefy_V4_reads_only, V4_meta)
 V1_alphadiv <- alpha_cal_df(rarefy_V1_reads_only, V1_meta)
 V6_alphadiv <- alpha_cal_df(rarefy_V6_reads_only, V6_meta)
 ITS1_alphadiv <- alpha_cal_df(rarefy_ITS1_reads_only, ITS1_meta)
+
+# pairwise t-test on richness and shannon
+library(rstatix)
+rich_pair_V4 <- pairwise_t_test(V4_alphadiv, S.obs ~ group)
+rich_pair_V1 <- pairwise_t_test(V1_alphadiv, S.obs ~ group)
+rich_pair_V6 <- pairwise_t_test(V6_alphadiv, S.obs ~ group)
+rich_pair_ITS1 <- pairwise_t_test(ITS1_alphadiv, S.obs ~ group)
+
+options(scipen = 10, digits = 2)
 
[…]

### 2024-02-03 analysis/paper/paper.qmd
work on alpha diversity

@@
 )
-
 ```
@@
 rarefy_ITS1 <- read.csv(here::here("analysis","data","derived_data", "ITS1_shared_format.1.subsample.387.csv"), sep = ",")
+context <- read_excel(here::here("analysis","data", "raw_data", "GJB_amp_meta.xlsx"))
 
@@
 
+```{r alpha}
+#| label: alpha
+#| echo: false
+#| cache: true
+library(vegan) 
+# to get the same order for all
+rarefy_V4 <- rarefy_V4 %>% mutate(Group = tolower(Group)) %>% arrange(Group)
+rarefy_V1 <- rarefy_V1 %>% mutate(Group = tolower(Group)) %>% arrange(Group)
+rarefy_V6 <- rarefy_V6 %>% mutate(Group = tolower(Group)) %>% arrange(Group)
+rarefy_ITS1 <- rarefy_ITS1 %>% mutate(Group = tolower(Group)) %>% arrange(Group)
+
+# tidy meta data and get the same order as the rarefy data
+context_tidy <-
+  context %>%
+  mutate(`adpater no` = tolower(`adpater no`),
+         type = tolower(str_extract(`adpater no`, "[[:alpha:]]+(?=\\d)")),
+         type = case_when(type == "con" ~ "control", TRUE ~ as.character(type)),
+         area = case_when(area == "inside" ~ "interior", area == "outside" ~ "exterior"),
+         group = if_else(!is.na(area)&!area%in% c("belly", "outside belly"),
+                         paste0(type, "-", area), type),
+         group = factor(group, levels = c("control","pot-interior","pot-exterior","soil"))) %>%
+  arrange(`adpater no`)
+
+# function to have OTU reads only
+reads_only <- function(df) {
+  df %>%
+    select(-label, -Group, -numOtus)
+}
+
+rarefy_V4_reads_only  <- reads_only(rarefy_V4)
+rarefy_V1_reads_only  <- reads_only(rarefy_V1)
+rarefy_V6_reads_only  <- reads_only(rarefy_V6)
+rarefy_ITS1_reads_only  <- reads_only(rarefy_ITS1)
+
+# function to get context for each region
+meta_region <- function(df, region){
+  context_tidy %>%
+    filter(Region == region) %>%
+    arrange(`adpater no`)
+}
+
+V4_meta<- meta_region(context_tidy, "V4")
+V1_meta<- meta_region(context_tidy, "V1")
+V6_meta<- meta_region(context_tidy, "V6")
+ITS1_meta<- meta_region(context_tidy, "ITS1")
+
+# functions calculate richness, Chao1, evenness index, and Shannon index 
+alpha_cal_df <- function(df1, df2){
+  richness <- estimateR(df1)
+  evenness <- diversity(df1)/log(specnumber(df1))
+  shannon <- diversity(df1, index = "shannon")
+  alphadiv <- cbind(df2, t(richness), shannon, evenness) #combine all in one
+  return(alphadiv)
+}
+
+V4_alphadiv <- alpha_cal_df(rarefy_V4_reads_only, V4_meta)
+V1_alphadiv <- alpha_cal_df(rarefy_V1_reads_only, V1_meta)
+V6_alphadiv <- alpha_cal_df(rarefy_V6_reads_only, V6_meta)
+ITS1_alphadiv <- alpha_cal_df(rarefy_ITS1_reads_only, ITS1_meta)
+
+# pairwise t-test on richness and shannon
+library(rstatix)
+rich_pair_V4 <- pairwise_t_test(V4_alphadiv, S.obs ~ group, p.adjust.method = 'none')
+rich_pair_V1 <- pairwise_t_test(V1_alphadiv, S.obs ~ group, p.adjust.method = 'none')
+rich_pair_V6 <- pairwise_t_test(V6_alphadiv, S.obs ~ group, p.adjust.method = 'none')
+rich_pair_ITS1 <- pairwise_t_test(ITS1_alphadiv, S.obs ~ group, p.adjust.method = 'none')
+
+shan_pair_V4 <- pairwise_t_test(V4_alphadiv, shannon ~ group, p.adjust.method = 'none')
+shan_pair_V1 <- pairwise_t_test(V1_alphadiv, shannon ~ group, p.adjust.method = 'none')
+shan_pair_V6 <- pairwise_t_test(V6_alphadiv, shannon ~ group, p.adjust.method = 'none')
+shan_pair_ITS1 <- pairwise_t_test(ITS1_alphadiv, shannon ~ group, p.adjust.method = 'none')
+
+rich_pv_v4 <- format.pval(pv = rich_pair_V4$p[6], 
+              digits = 2,  # number of digits, but after the 0.0
+              eps = 0.001, # above the number will be replaced  by "<0.0xxx"
+              nsmall = 3) # how much tails 0 to keep
+
+rich_pv_v1 <- format.pval(pv = rich_pair_V1$p[4], digits = 2, eps = 0.01, nsmall = 2) 
+rich_pv_its1 <- format.pval(pv = rich_pair_ITS1$p[4], digits = 2, eps = 0.05, nsmall = 2)
+shan_pv_v4 <- format.pval(pv = shan_pair_V4$p[2], digits = 2, eps = 0.05, nsmall = 2) 
+shan_pv_v1 <- format.pval(pv = shan_pair_V1$p[2], digits = 2, eps = 0.05, nsmall = 2) 
+```
+
 ## Alpha diversity analysis of bacterial and fungal communities
-Illumina analysis generated a total of `r total_reads` sequences that were normalized to `r rare_reads` after rarefaction [@Schloss2024].
-Soil samples have the highest shannon diversity index and richness, followed by pottery exterior samples, and pottery interior samples, and negative control.
+Illumina analysis generated a total of `r total_reads` sequences that were normalized to `r rare_reads` after rarefaction for comparison across different types of samples [@Schloss2024].
+For the V4 and V6 regions, soil samples show a significant higher richness indices based on the number of observed OTUs (p `r rich_pv_v4`) compared with those of exterior pottery samples, interior pottery samples, and negative control samples.
+Similar patterns were observed for the V1 and ITS1 region, but the richness index differ significantly mainly between soil samples and negative control samples (p `r rich_pv_v1` and `r rich_pv_its1` respectively; Table S1).
+Although there is no statistically significant differences in richness indices between the two types of pottery samples and negative control, Shannon diversity indices for the V4 and V1 regions are significantly higher from exterior pottery samples than those from negative control samples (p `r shan_pv_v4` and `r shan_pv_v1` respectively; Table S2).
 

### 2024-02-03 analysis/supplementary-materials/Supplement.qmd
create supplementary material

---
title: "Supplementary Materials for Characterization of the microbial and fungal communities in ancient pottery from Neolithic southwestern China"
author:
  - Li-Ying Wang:
      correspondence: "yes"
      email: liying15@gmail.com
      orcid: 0000-0001-9060-2125
      institute:
        - AS
  - Kuei-Chen Lin:
      institute: AS
      orcid: 0000-0003-1689-0558
  - Zhiqing Zhou:
      orcid: 0000-0003-1689-0551
      institute:
        - name: Chengdu Institute of Cultural Relics and Archaeology
  - Yin-Ru Chiang:
      correspondence: "yes"
      institute: AS
      orcid: 0000-0003-1689-0559
institute:
  - AS:
      name: Academia sinica
      address: 130, Section 2, Academia Rd, Nangang District, Taipei City, Taiwan
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
---

## Alpha Diversity

```{r get-data}
#| label: get-data
#| echo: false
#| cache: true
library(tidyverse)
library(readxl)
# read rarefied data
rarefy_V4 <- read.csv(here::here("analysis","data","derived_data", "V4_shared_format.1.subsample.60216.csv"), sep = ",")
rarefy_V1 <- read.csv(here::here("analysis","data","derived_data","V1_shared_format.1.subsample.13658.csv"), sep = "")
rarefy_V6 <- read.csv(here::here("analysis","data","derived_data", "V6_shared_format.1.subsample.28402.csv"), sep = ",")
rarefy_ITS1 <- read.csv(here::here("analysis","data","derived_data", "ITS1_shared_format.1.subsample.387.csv"), sep = ",")
context <- read_excel(here::here("analysis","data", "raw_data", "GJB_amp_meta.xlsx"))
```

```{r}
#| label: alpha-richness
#| echo: false
#| cache: true
library(knitr)
library(vegan) 
# to get the same order for all
rarefy_V4 <- rarefy_V4 %>% mutate(Group = tolower(Group)) %>% arrange(Group)
rarefy_V1 <- rarefy_V1 %>% mutate(Group = tolower(Group)) %>% arrange(Group)
rarefy_V6 <- rarefy_V6 %>% mutate(Group = tolower(Group)) %>% arrange(Group)
rarefy_ITS1 <- rarefy_ITS1 %>% mutate(Group = tolower(Group)) %>% arrange(Group)

# tidy meta data and get the same order as the rarefy data
context_tidy <-
  context %>%
  mutate(`adpater no` = tolower(`adpater no`),
         type = tolower(str_extract(`adpater no`, "[[:alpha:]]+(?=\\d)")),
         type = case_when(type == "con" ~ "control", TRUE ~ as.character(type)),
         area = case_when(area == "inside" ~ "interior", area == "outside" ~ "exterior"),
         group = if_else(!is.na(area)&!area%in% c("belly", "outside belly"),
                         paste0(type, "-", area), type),
         group = factor(group, levels = c("control","pot-interior","pot-exterior","soil"))) %>%
  arrange(`adpater no`)

# function to have OTU reads only
reads_only <- function(df) {
  df %>%
    select(-label, -Group, -numOtus)
}

rarefy_V4_reads_only  <- reads_only(rarefy_V4)
rarefy_V1_reads_only  <- reads_only(rarefy_V1)
rarefy_V6_reads_only  <- reads_only(rarefy_V6)
rarefy_ITS1_reads_only  <- reads_only(rarefy_ITS1)

# function to get context for each region
meta_region <- function(df, region){
  context_tidy %>%
    filter(Region == region) %>%
    arrange(`adpater no`)
}

V4_meta<- meta_region(context_tidy, "V4")
V1_meta<- meta_region(context_tidy, "V1")
V6_meta<- meta_region(context_tidy, "V6")
ITS1_meta<- meta_region(context_tidy, "ITS1")

# functions calculate richness, Chao1, evenness index, and Shannon index 
alpha_cal_df <- function(df1, df2){
  richness <- estimateR(df1)
  evenness <- diversity(df1)/log(specnumber(df1))
  shannon <- diversity(df1, index = "shannon")
  alphadiv <- cbind(df2, t(richness), shannon, evenness) #combine all in one
  return(alphadiv)
}

V4_alphadiv <- alpha_cal_df(rarefy_V4_reads_only, V4_meta)
V1_alphadiv <- alpha_cal_df(rarefy_V1_reads_only, V1_meta)
V6_alphadiv <- alpha_cal_df(rarefy_V6_reads_only, V6_meta)
ITS1_alphadiv <- alpha_cal_df(rarefy_ITS1_reads_only, ITS1_meta)

# pairwise t-test on richness
library(rstatix)
# no correction for the Type I error rate across the pairwise tests
rich_pair_V4 <- pairwise_t_test(V4_alphadiv, S.obs ~ group, p.adjust.method = 'none')
rich_pair_V1 <- pairwise_t_test(V1_alphadiv, S.obs ~ group, p.adjust.method = 'none')
rich_pair_V6 <- pairwise_t_test(V6_alphadiv, S.obs ~ group, p.adjust.method = 'none')
rich_pair_ITS1 <- pairwise_t_test(ITS1_alphadiv, S.obs ~ group, p.adjust.method = 'none')

# function to add region name
rich_pair_region <- function(df, vector) {
  df %>%
    mutate(region = vector) %>% 
    select(region, group1, group2, n1, n2, p, p.signif)
}

rich_pair_V4 <- rich_pair_region(rich_pair_V4, "V4")
rich_pair_V1 <- rich_pair_region(rich_pair_V1, "V1")
rich_pair_V6 <- rich_pair_region(rich_pair_V6, "V6")
rich_pair_ITS1 <- rich_pair_region(rich_pair_ITS1, "ITS1")

rich_pair_all <-
  rbind(rich_pair_V1, rich_pair_V4, rich_pair_V6, rich_pair_ITS1)

kable(rich_pair_all , caption = "Table S1: Pairwise comparison of Richness index (the number of observed OTUs) using ANOVA")
```

```{r}
#| label: alpha-shannon
#| echo: false
#| cache: true
# pairwise t-test on Shannon

shan_pair_V4 <- pairwise_t_test(V4_alphadiv, shannon ~ group, p.adjust.method = 'none')
shan_pair_V1 <- pairwise_t_test(V1_alphadiv, shannon ~ group, p.adjust.method = 'none')
shan_pair_V6 <- pairwise_t_test(V6_alphadiv, shannon ~ group, p.adjust.method = 'none')
shan_pair_ITS1 <- pairwise_t_test(ITS1_alphadiv, shannon ~ group, p.adjust.method = 'none')

# function to add region name
shan_pair_region <- function(df, vector) {
  df %>%
    mutate(region = vector) %>% 
    select(region, group1, group2, n1, n2, p, p.signif)
}

shan_pair_V4 <- shan_pair_region(shan_pair_V4, "V4")
shan_pair_V1 <- shan_pair_region(shan_pair_V1, "V1")
shan_pair_V6 <- shan_pair_region(shan_pair_V6, "V6")
shan_pair_ITS1 <- shan_pair_region(shan_pair_ITS1, "ITS1")

shan_pair_all <-
  rbind(shan_pair_V1, shan_pair_V4, shan_pair_V6, shan_pair_ITS1)

kable(shan_pair_all  , caption = "Table S2: Pairwise comparison of Shannon diversity indices using ANOVA")

```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2024-02-05 004-taxa-relative-ab.R
export rarefied data

@@ -56,6 +56,7 @@ V4_taxa_percent_group <-
   ungroup()
 
 # Archaeal only: relative abundance plot (the top ones with others)
+library(RColorBrewer)
 V6_archaeal_count_ab  <-
   V6_tax_rarefy_df %>%
   filter(Domain == "k__Archaea") %>%
@@ -101,18 +102,17 @@ ITS1_Class_group <- ITS1_tax_rarefy_df %>%
   arrange(Class) %>% #in a alphabetic order
   pull()
 
-library(RColorBrewer)
 phylum_colors_rel_mod<- c("#D9D9D9","#FFFF99","#B2DF8A","#33A02C","#1F78B4","#E31A1C", # "C51B7D"
                           "#FDBF6F","#FF7F00","#CAB2D6","#6A3D9A","#B15928","#FB9A99") #brewer.pal(12,"Paired")
 phylum_colors_rel_mod2<- c("#D9D9D9","#FFFF99","#B2DF8A","#F781BF","#1F78B4","#E31A1C",
-                           "#FF7F00","#CAB2D6","#6A3D9A","#B15928","#A6CEE3","#FB9A99")
+                           "#FF7F00","#CAB2D6","#6A3D9A","#B15928","#A6CEE3","#FB9A99") # for V6
 phylum_colors_rel_mod3<- c("#D9D9D9","#FDAE61","#FFED6F","#BC80BD","#80B1D3","#D6604D",
-                           "#B3DE69","#8DD3C7","#BEBADA","#CCEBC5","#FCCDE5","#DFC27D")
+                           "#B3DE69","#8DD3C7","#BEBADA","#CCEBC5","#FCCDE5","#DFC27D") # for ITS1
 
 # V4 relative abundance plot (top ones are others and unclassified)
-V4_bac_phyla_rel_ab <-
-  V4_tax_rarefy_df %>%
-  mutate(Phylum = ifelse(!Phylum%in% V4_Phylum_group, "others", Phylum)) %>%
+V6_bac_phyla_rel_ab <- # region number can be replaced
+  V6_tax_rarefy_df %>%
+  mutate(Phylum = ifelse(!Phylum%in% V6_Phylum_group, "others", Phylum)) %>%
   mutate(Taxonomy = ifelse(Domain == "unclassified" & Phylum == "unclassified"|
                            Domain == "k__Bacteria"& Phylum == "unclassified"|
[…]

### 2024-02-05 analysis/supplementary-materials/Supplement.qmd
work on supplement

@@
 
-# pairwise t-test on richness
+# pairwise Wilcoxon rank-sum  on richness and shannon
 library(rstatix)
-# no correction for the Type I error rate across the pairwise tests
-rich_pair_V4 <- pairwise_t_test(V4_alphadiv, S.obs ~ group, p.adjust.method = 'none')
-rich_pair_V1 <- pairwise_t_test(V1_alphadiv, S.obs ~ group, p.adjust.method = 'none')
-rich_pair_V6 <- pairwise_t_test(V6_alphadiv, S.obs ~ group, p.adjust.method = 'none')
-rich_pair_ITS1 <- pairwise_t_test(ITS1_alphadiv, S.obs ~ group, p.adjust.method = 'none')
+
+rich_pair_V4 <- V4_alphadiv %>% pairwise_wilcox_test(S.obs ~ group, p.adjust.method = 'fdr')
+rich_pair_V1 <- V1_alphadiv %>% pairwise_wilcox_test(S.obs ~ group, p.adjust.method = 'fdr')
+rich_pair_V6 <- V6_alphadiv %>% pairwise_wilcox_test(S.obs ~ group, p.adjust.method = 'fdr')
+rich_pair_ITS1 <- ITS1_alphadiv %>% pairwise_wilcox_test(S.obs ~ group, p.adjust.method = 'fdr')
 
@@
     mutate(region = vector) %>% 
-    select(region, group1, group2, n1, n2, p, p.signif)
+    select(region, group1, group2, n1, n2, p, p.adj, p.adj.signif)
 }
@@
 
-kable(rich_pair_all , caption = "Table S1: Pairwise comparison of Richness index (the number of observed OTUs) using ANOVA")
+kable(rich_pair_all , caption = "Table S1: Pairwise comparison of Richness index (the number of observed OTUs) using Wilcoxon rank-sum test with FDR-adjusted p-value")
 ```
@@
 #| cache: true
+
 # pairwise t-test on Shannon
-
-shan_pair_V4 <- pairwise_t_test(V4_alphadiv, shannon ~ group, p.adjust.method = 'none')
-shan_pair_V1 <- pairwise_t_test(V1_alphadiv, shannon ~ group, p.adjust.method = 'none')
-shan_pair_V6 <- pairwise_t_test(V6_alphadiv, shannon ~ group, p.adjust.method = 'none')
-shan_pair_ITS1 <- pairwise_t_test(ITS1_alphadiv, shannon ~ group, p.adjust.method = 'none')
+shan_pair_V4 <- V4_alphadiv %>% pairwise_wilcox_test(shannon ~ group, p.adjust.method = 'fdr')
+shan_pair_V1 <- V1_alphadiv %>% pairwise_wilcox_test(shannon ~ group, p.adjust.method = 'fdr')
+shan_pair_V6 <- V6_alphadiv %>% pairwise_wilcox_test(shannon ~ group, p.adjust.method = 'fdr')
+shan_pair_ITS1 <- ITS1_alphadiv %>% pairwise_wilcox_test(shannon ~ group, p.adjust.method = 'fdr')
 
@@
 
-kable(shan_pair_all  , caption = "Table S2: Pairwise comparison of Shannon diversity indices using ANOVA")
-
+kable(shan_pair_all  , caption = "Table S2: Pairwise comparison of Shannon diversity indices using Wilcoxon rank-sum test with FDR-adjusted p-value")
 ```

### 2024-02-05 000-tidy-data.R
edit code for label

@@ -20,18 +20,22 @@ context <- context %>%
                          paste0(type, "-", area), type),
          group = case_when(group == "pot-inside"~ "pot-interior",
                            group == "pot-outside"~ "pot-exterior", TRUE ~ as.character(group))) %>%
-  mutate(label = case_when(`adpater no` == "Pot1"~ "pot1\nin", `adpater no` == "Pot2"~ "pot2\nin",
-                           `adpater no` == "Pot3"~ "pot3\nin", `adpater no` == "Pot4"~ "pot4\nin",
-                           `adpater no` == "Pot5"~ "pot5\nin", `adpater no` == "Pot11"~ "pot6\nin",
-                           `adpater no` == "Pot6"~ "pot1\nex", `adpater no` == "Pot7"~ "pot2\nex",
-                           `adpater no` == "Pot8"~ "pot3\nex", `adpater no` == "Pot9"~ "pot4\nex",
-                           `adpater no` == "Pot10"~ "pot5\nex", `adpater no` == "Pot12"~ "pot6\nex",
-                           `adpater no` == "soil4"~ "soil4\nbelly", `adpater no` == "soil5"~ "soil5\nbelly\nex",
+  mutate(label = case_when(`adpater no` == "con1"~ "C1\nblank", `adpater no` == "con2"~ "C2\nblank",
+                           `adpater no` == "Pot1"~ "P1\nGJB", `adpater no` == "Pot2"~ "P2\nGJB",
+                           `adpater no` == "Pot3"~ "P3\nGJB", `adpater no` == "Pot4"~ "P4\nGJB",
+                           `adpater no` == "Pot5"~ "P5\nGJB", `adpater no` == "Pot11"~ "P6\nGS",
+                           `adpater no` == "Pot6"~ "P1\nGJB", `adpater no` == "Pot7"~ "P2\nGJB",
+                           `adpater no` == "Pot8"~ "P3\nGJB", `adpater no` == "Pot9"~ "P4\nGJB",
+                           `adpater no` == "Pot10"~ "P5\nGJB", `adpater no` == "Pot12"~ "P6\nGS",
+                           `adpater no` == "soil1"~ "S1\nGJB", `adpater no` == "soil2"~ "S2\nGJB",
+                           `adpater no` == "soil3"~ "S3\nGJB", `adpater no` == "con3"~ "C3\nPCR",
[…]

### 2024-02-05 004-taxa-relative-ab.R
revise the relative abundance plot

@@ -31,7 +31,7 @@ tax_rarefy_df <- function(df1, df2, df3){
     select(-label, -numOtus) %>%
     left_join(df2, by = c("Group"= "adpater no")) %>% #V4_context
     arrange(type, part, order) %>%
-    mutate(label = factor(label, levels = label)) %>%
+    #mutate(label = factor(label, levels = label)) %>%
     select(-Group, -area, -Region) %>%
     pivot_longer(!c(sample, label, type, group, site, part, order), names_to = "name", values_to = "count") %>%
     left_join(df3, by = c("name"= "OTU.ID")) %>% #V4_tax_clean_sep
@@ -110,9 +110,10 @@ phylum_colors_rel_mod3<- c("#D9D9D9","#FDAE61","#FFED6F","#BC80BD","#80B1D3","#D
                            "#B3DE69","#8DD3C7","#BEBADA","#CCEBC5","#FCCDE5","#DFC27D") # for ITS1
 
 # V4 relative abundance plot (top ones are others and unclassified)
-V6_bac_phyla_rel_ab <- # region number can be replaced
-  V6_tax_rarefy_df %>%
-  mutate(Phylum = ifelse(!Phylum%in% V6_Phylum_group, "others", Phylum)) %>%
+V4_bac_phyla_rel_ab <- # region number can be replaced
+  V4_tax_rarefy_df %>%
+  #mutate(type = case_when(type == "con" ~ "control", type == "pot" ~ "pottery", TRUE ~ "Other"))
+  mutate(Phylum = ifelse(!Phylum%in% V4_Phylum_group, "others", Phylum)) %>%
   mutate(Taxonomy = ifelse(Domain == "unclassified" & Phylum == "unclassified"|
                            Domain == "k__Bacteria"& Phylum == "unclassified"|
                            Domain == "k__Archaea" & Phylum == "unclassified",
@@ -121,11 +122,13 @@ V6_bac_phyla_rel_ab <- # region number can be replaced
                               Taxonomy == "k__Archaea; others" ~ "others", TRUE ~ as.character(Taxonomy))) %>%
   ggplot(aes(x = label, y = count, fill = Taxonomy)) +
   geom_bar(stat="identity", position = position_fill(reverse = TRUE), width = 0.8) + #"stack" shows original count; # fill
-  scale_y_continuous(labels = scales::percent) +
[…]

### 2024-02-05 analysis/paper/paper.qmd
work on the result section

@@
 
-# pairwise t-test on richness and shannon
+# pairwise Wilcoxon rank-sum  on richness and shannon
 library(rstatix)
-rich_pair_V4 <- pairwise_t_test(V4_alphadiv, S.obs ~ group, p.adjust.method = 'none')
-rich_pair_V1 <- pairwise_t_test(V1_alphadiv, S.obs ~ group, p.adjust.method = 'none')
-rich_pair_V6 <- pairwise_t_test(V6_alphadiv, S.obs ~ group, p.adjust.method = 'none')
-rich_pair_ITS1 <- pairwise_t_test(ITS1_alphadiv, S.obs ~ group, p.adjust.method = 'none')
+rich_pair_V4 <- V4_alphadiv %>% pairwise_wilcox_test(S.obs ~ group, p.adjust.method = 'fdr')
+rich_pair_V1 <- V1_alphadiv %>% pairwise_wilcox_test(S.obs ~ group, p.adjust.method = 'fdr')
+rich_pair_V6 <- V6_alphadiv %>% pairwise_wilcox_test(S.obs ~ group, p.adjust.method = 'fdr')
+rich_pair_ITS1 <- ITS1_alphadiv %>% pairwise_wilcox_test(S.obs ~ group, p.adjust.method = 'fdr')
 
-shan_pair_V4 <- pairwise_t_test(V4_alphadiv, shannon ~ group, p.adjust.method = 'none')
-shan_pair_V1 <- pairwise_t_test(V1_alphadiv, shannon ~ group, p.adjust.method = 'none')
-shan_pair_V6 <- pairwise_t_test(V6_alphadiv, shannon ~ group, p.adjust.method = 'none')
-shan_pair_ITS1 <- pairwise_t_test(ITS1_alphadiv, shannon ~ group, p.adjust.method = 'none')
+shan_pair_V4 <- V4_alphadiv %>% pairwise_wilcox_test(shannon ~ group, p.adjust.method = 'fdr')
+shan_pair_V1 <- V1_alphadiv %>% pairwise_wilcox_test(shannon ~ group, p.adjust.method = 'fdr')
+shan_pair_V6 <- V6_alphadiv %>% pairwise_wilcox_test(shannon ~ group, p.adjust.method = 'fdr')
+shan_pair_ITS1 <- ITS1_alphadiv %>% pairwise_wilcox_test(shannon ~ group, p.adjust.method = 'fdr')
+rich_pv_v4 <- format.pval(pv = rich_pair_V4$p.adj[6], 
+              digits = 2,  # number of digits, but after the 0.0
+              eps = 0.05, # above the number will be replaced  by "<0.0xxx"
+              nsmall = 2) # how much tails 0 to keep
 
-rich_pv_v4 <- format.pval(pv = rich_pair_V4$p[6], 
-              digits = 2,  # number of digits, but after the 0.0
-              eps = 0.001, # above the number will be replaced  by "<0.0xxx"
-              nsmall = 3) # how much tails 0 to keep
-
-rich_pv_v1 <- format.pval(pv = rich_pair_V1$p[4], digits = 2, eps = 0.01, nsmall = 2) 
-rich_pv_its1 <- format.pval(pv = rich_pair_ITS1$p[4], digits = 2, eps = 0.05, nsmall = 2)
-shan_pv_v4 <- format.pval(pv = shan_pair_V4$p[2], digits = 2, eps = 0.05, nsmall = 2) 
-shan_pv_v1 <- format.pval(pv = shan_pair_V1$p[2], digits = 2, eps = 0.05, nsmall = 2) 
+rich_pv_v6 <- format.pval(pv = rich_pair_V6$p.adj[2], digits = 2, eps = 0.05, nsmall = 2) 
+shan_pv_v4 <- format.pval(pv = shan_pair_V4$p[5], digits = 2, eps = 0.05, nsmall = 2) 
 ```
@@
 ## Alpha diversity analysis of bacterial and fungal communities
-Illumina analysis generated a total of `r total_reads` sequences that were normalized to `r rare_reads` after rarefaction for comparison across different types of samples [@Schloss2024].
-For the V4 and V6 regions, soil samples show a significant higher richness indices based on the number of observed OTUs (p `r rich_pv_v4`) compared with those of exterior pottery samples, interior pottery samples, and negative control samples.
-Similar patterns were observed for the V1 and ITS1 region, but the richness index differ significantly mainly between soil samples and negative control samples (p `r rich_pv_v1` and `r rich_pv_its1` respectively; Table S1).
-Although there is no statistically significant differences in richness indices between the two types of pottery samples and negative control, Shannon diversity indices for the V4 and V1 regions are significantly higher from exterior pottery samples than those from negative control samples (p `r shan_pv_v4` and `r shan_pv_v1` respectively; Table S2).
+Illumina analysis generated a total of `r total_reads` sequences that were normalized to `r rare_reads` after rarefying for comparison across different types of samples [@Schloss2024; @Weiss2017].
+Soil samples show a significant higher richness index based on the number of observed OTUs than that of pottery exterior samples and pottery interior samples (p `r rich_pv_v4`) for the the V4 and V6 regions (Table S1).
+Also, the richness index of pottery exterior samples is significant higher than that of control samples for the V6 region (p `r rich_pv_v6`).
+The richness index did not differ significantly between samples for the V1 and ITS1 regions.
+Shannon diversity index shows that soil samples are significantly higher than pottery interior samples (p `r shan_pv_v4`; Table S2).
+However, there is no statistically significant differences in richness index and Shannon diversity index between the two types of pottery samples and negative control.
+
+```{r taxa-abundance}
+#| label: taxa-abundance
+#| echo: false
+#| cache: true
+V4_taxa_ab <- read.csv(here::here("analysis","data","derived_data","V4_bac_phyla_rel_ab.csv"))
+V6_taxa_ab <- read.csv(here::here("analysis","data","derived_data","V6_bac_phyla_rel_ab.csv"))
+V4_taxa_all_sample <- 
+  V4_taxa_ab %>%   # OTU_df_V4_original should be ran in 000-tidy-data
+  group_by(label) %>% 
+  mutate(total_phylum_sum = sum(count)) %>% 
+  ungroup() %>% 
+  group_by(Phylum, label) %>% 
+  mutate(each_phylum_sum = sum(count)) %>% 
+  ungroup() %>% 
+  distinct(label, Phylum, total_phylum_sum, each_phylum_sum) %>% 
+  mutate(rel_ab = round(100*each_phylum_sum/total_phylum_sum, 1))
+
+pottery_GJB <-V4_taxa_all_sample %>% filter(!str_detect(label, "pot6")&!str_detect(label, "soil")&!str_detect(label, "con"))
+pottery_GS <-V4_taxa_all_sample %>% filter(str_detect(label, "pot6"))
+Firmi <- pottery_GJB %>% filter(Phylum == "p__Firmicutes")
+Actin <- pottery_GJB %>% filter(Phylum == "p__Actinobacteria")
+Deino <- pottery_GJB %>% filter(Phylum == "p__Deinococcus")
+Prote <- pottery_GJB %>% filter(Phylum == "p__Proteobacteria")
+pot6_Actin <- pottery_GS %>% filter(Phylum == "p__Actinobacteria")
+pot6_Crena <- pottery_GS %>% filter(Phylum == "p__Crenarchaeota")
+
+soil <-V4_taxa_all_sample %>% filter(str_detect(label, "soil"))
+Firmi_soil <- soil %>% filter(Phylum == "p__Firmicutes"& label%in%c("soil1", "soil2", "soil3"))
+Thau_soil <- soil %>% filter(Phylum == "p__Thaumarchaeota"& label%in%c("soil1", "soil2", "soil3"))
+Actin_soil4 <- soil %>% filter(str_detect(label, "soil4") & Phylum == "p__Actinobacteria")
+Actin_soil5 <- soil %>% filter(str_detect(label, "soil5") & Phylum == "p__Actinobacteria")
+Thau_soil4 <- soil %>% filter(str_detect(label, "soil4") & Phylum == "p__Thaumarchaeota")
+Thau_soil5 <- soil %>% filter(str_detect(label, "soil5") & Phylum == "p__Thaumarchaeota")
+
+library(corrr)
+rel_ab_wider <- 
+  V4_taxa_all_sample %>% 
+  select(label, Phylum, each_phylum_sum) %>% #each_phylum_sum, rel_ab
+  pivot_wider(names_from = label, values_from = each_phylum_sum) %>%  
+  select(-Phylum)
+
+# compute correlation test
+cor_rel_ab <- rel_ab_wider %>%  correlate()
+
+# Pearson correlation test 
+calc_cor_p_value <- function(vec_a, vec_b){
+  cor.test(vec_a, vec_b)$p.value
+}
+p_value_cor <- colpair_map(rel_ab_wider, calc_cor_p_value)
+
+# Wilcoxon rank-sum
+calc_wil_p_value <- function(vec_a, vec_b){
+  wilcox.test(vec_a, vec_b, p.adjust.method = 'fdr')$p.value
+}
+p_value_wil<- colpair_map(rel_ab_wider, calc_wil_p_value)
+
+# GLM
+calc_glm_p_value <- function(vec_a, vec_b){
+  x <- glm(vec_a ~ vec_b, family = poisson)
+  broom::tidy(x)$p.value[2]
+}
+p_value_glm <- colpair_map(rel_ab_wider, calc_glm_p_value)
+
+major_taxa <- function(df, vec1){
+  df %>% 
+  filter(Phylum == vec1) %>% 
+  group_by(label) %>% 
+  mutate(total_sum = sum(count)) %>% 
+  ungroup() %>% 
+  group_by(Class, label) %>% 
+  mutate(each_sum = sum(count)) %>% 
+  ungroup() %>% 
+  distinct(label, Class, total_sum, each_sum) %>% 
+  mutate(rel_ab = round(100*each_sum/total_sum, 1))
+}
+
+V4_Firmi <- major_taxa(V4_taxa_ab, "p__Firmicutes")
+
+Bacilli <- V4_Firmi %>% filter(Class == "c__Bacilli")
+Clost <- V4_Firmi %>% filter(Class == "c__Clostridia")
+Nega <- V4_Firmi %>% filter(Class == "c__Negativicutes")
+
+V4_Actin <- major_taxa(V4_taxa_ab, "p__Actinobacteria")
+V4_Deino <- major_taxa(V4_taxa_ab, "p__Deinococcus")
+V4_Prote <- major_taxa(V4_taxa_ab, "p__Proteobacteria")
+
+```
 
 ## Taxonomic composition 
+For the V4 region, bacterial composition at the phylum level in pottery interior samples was similar to those of pottery exterior samples and control samples, but slightly different from that of soil samples (@fig-V4-rel-ab).
+For Guijiabao pottery samples, *Firmicutes* (`r min(Firmi$rel_ab)`- `r max(Firmi$rel_ab)`%) was the most abundant phylum, followed by *Actinobacteria* (`r min(Actin$rel_ab)`- `r max(Actin$rel_ab)`%), *Deinococcus* (`r min(Deino$rel_ab)`- `r max(Deino$rel_ab)`%), and *Proteobacteria* (`r min(Prote$rel_ab)`- `r max(Prote$rel_ab)`%).
+Gaoshan sample is predominated by *Actinobacteria* (`r min(pot6_Actin$rel_ab)`- `r max(pot6_Actin$rel_ab)`%) with the presence of *Crenarchaeota* (`r min(pot6_Crena$rel_ab)`- `r max(pot6_Crena$rel_ab)`%) belonging to archaeal domain that was not found in Guijiabao samples and control samples.
+Although a different composition of phylum was observed between Guijiabao and Gaoshan samples, the difference is not statistically significant.
+The soil samples from Guijiabao is predominated by *Firmicutes* (`r min(Firmi_soil$rel_ab)`- `r max(Firmi_soil$rel_ab)`%) and *Thaumarchaeota* (`r min(Thau_soil$rel_ab)`- `r max(Thau_soil$rel_ab)`%) belonging to the archaeal domain, while soil samples from Gaoshan is predominated by *Actinobacteria* (`r Actin_soil4$rel_ab`- `r Actin_soil5$rel_ab`%) with less abundant *Thaumarchaeota* (`r Thau_soil4$rel_ab`- `r Thau_soil5$rel_ab`%).
+The dominant classes include *Bacilli*, *Clostridia* and *Negativicutes* among *Firmicutes*; *Actinobacteria*, *Rubrobacteria*, *Thermoleophilia* among *Actinobacteria*, *Deinococci* in *Deinococcus*, and *Gammaproteobacteria*, *Alphaproteobacteria*, and *Deltaproteobacteria* among *Proteobacteria*.
+
+```{r}
+#| label: fig-V4-rel-ab
+#| echo: false
+#| cache: true
+#| fig-cap: "The taxonomic composition of bacterial communities for the V4 region of pottery, soil, and control samples at the genus level."
+library(here)
+knitr::include_graphics(here("analysis", "figures", "03_rel_ab_V4.png"))
+```
 
@@
 Kristall_willis
-
-<!-- Here's some example analysis code: -->
-
-```{r}
-#| label: fig-demo-plot
-#| fig-cap: "A plot of random numbers"
-plot(rnorm(10))
-```
-
-@fig-demo-plot shows how we can have a caption and cross-reference for a plot. Note that figure label and cross-references must both be prefixed with `fig-`
-
-```{r}
-#| label: demo-inline-code
-#| echo: false
-x <- round(pi, 2)
-```
-
-Here is an example of inline code `r x` in the middle of a sentence.
 

### 2024-02-07 003-rarefy-taxa-test.R
edit code

@@ -48,6 +48,7 @@ meta_region_extract <- function(filename, vec){
            pot = case_when(group == "pot-inside" ~ "interior", group == "pot-outside" ~"exterior"),
            potex_soil = case_when(group == "pot-outside" ~ "pot-exterior", type== "soil" ~ "soil"),
            potin_soil = case_when(group == "pot-inside" ~ "pot-interior", type== "soil" ~ "soil"),
+           pot_soil = case_when(type== "pot" ~ "pot", type== "soil" ~ "soil"),
            pot_con = case_when(type== "pot" ~ "pot", type== "con" ~ "control"),
            potex_con = case_when(group == "pot-outside" ~ "pot-exterior", type== "con" ~ "control"),
            potin_con = case_when(group == "pot-inside" ~ "pot-interior", type== "con" ~ "control")) %>%
@@ -81,11 +82,11 @@ composite_ITS1 <- composite_region(shared_ITS1_rarefy, taxonomy_ITS1, metadata_I
 sig_genera_V4 <-
   composite_V4 %>%
   nest(data = -taxonomy) %>% #c(-taxonomy, -otu)
-  mutate(test = map(.x= data, ~wilcox.test(rel_abund~  pot_con, data=.x, exact = FALSE) %>% tidy)) %>% #run broom first
+  mutate(test = map(.x= data, ~wilcox.test(rel_abund~ potin_con, data=.x, exact = FALSE) %>% tidy)) %>% #run broom first
   unnest(test) %>%
-  filter(p.value < 0.1)
-  #mutate(p.adjust = p.adjust(p.value, method="BH")) %>% #https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6099145/
-  #filter(p.adjust < 0.01) %>%
+  filter(p.value < 0.05)
+  #mutate(p.adjust = p.adjust(p.value, method="fdr")) %>% #https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6099145/
+  #filter(p.adjust < 0.1)
   #select(taxonomy, p.adjust)
 
 sig_genera_V1 <-
@@ -145,12 +146,131 @@ composite_V4_plot <-
   theme(axis.text.y = element_markdown())
 
 # pairwise wilcox test
-pairwise.wilcox.test(composite_V4$rel_abund, composite_V4$group, p.adjust.method = "BH")
[…]

### 2024-02-09 analysis/paper/paper.qmd
work on results

@@
 
+At the family level, significant differences were observed between pottery exterior samples and control samples, and between pottery exterior and interior samples.
+*Paenibacillaceae*, *Erysipelotrichaceae*, and *Xiphinematobacteraceae* were more abundant in pottery exterior samples than those in control samples (P <0.05).
+In addition to *Paenibacillaceae* and *Xiphinematobacteraceae*, *Phycisphaeraceae* was more abundant in pottery exterior samples than in pottery interior samples (P <0.05).
+
 ```{r}

### 2024-02-10 analysis/paper/paper.qmd
work on relative abundance

@@
 The soil samples from Guijiabao is predominated by *Firmicutes* (`r min(Firmi_soil$rel_ab)`- `r max(Firmi_soil$rel_ab)`%) and *Thaumarchaeota* (`r min(Thau_soil$rel_ab)`- `r max(Thau_soil$rel_ab)`%) belonging to the archaeal domain, while soil samples from Gaoshan is predominated by *Actinobacteria* (`r Actin_soil4$rel_ab`- `r Actin_soil5$rel_ab`%) with less abundant *Thaumarchaeota* (`r Thau_soil4$rel_ab`- `r Thau_soil5$rel_ab`%).
-The dominant classes include *Bacilli*, *Clostridia* and *Negativicutes* among *Firmicutes*; *Actinobacteria*, *Rubrobacteria*, *Thermoleophilia* among *Actinobacteria*, *Deinococci* in *Deinococcus*, and *Gammaproteobacteria*, *Alphaproteobacteria*, and *Deltaproteobacteria* among *Proteobacteria*.
 
+The dominant classes include *Bacilli*, *Clostridia*, and *Negativicutes* among *Firmicutes*; *Actinobacteria*, *Rubrobacteria*, *Thermoleophilia* among *Actinobacteria*, *Deinococci* in *Deinococcus*, and *Gammaproteobacteria*, *Alphaproteobacteria*, and *Deltaproteobacteria* among *Proteobacteria*.
 At the family level, significant differences were observed between pottery exterior samples and control samples, and between pottery exterior and interior samples.
@@
 In addition to *Paenibacillaceae* and *Xiphinematobacteraceae*, *Phycisphaeraceae* was more abundant in pottery exterior samples than in pottery interior samples (P <0.05).
+At the genus level, pottery exterior samples exhibit significantly higher abundance of *Candidatus_Xiphinematobacter* compared to pottery interior samples and control samples (P < 0.01).
+Among the bacterial taxa showing significant differences, only *Paenibacillaceae* displayed relative abundances exceeding 10% on average, while others were below 1%.
 

### 2024-02-11 003-rarefy-taxa-test.R
edit plot

@@ -82,7 +82,7 @@ composite_ITS1 <- composite_region(shared_ITS1_rarefy, taxonomy_ITS1, metadata_I
 sig_genera_V4 <-
   composite_V4 %>%
   nest(data = -taxonomy) %>% #c(-taxonomy, -otu)
-  mutate(test = map(.x= data, ~wilcox.test(rel_abund~ potin_con, data=.x, exact = FALSE) %>% tidy)) %>% #run broom first
+  mutate(test = map(.x= data, ~wilcox.test(rel_abund~ pot, data=.x, exact = FALSE) %>% tidy)) %>% #run broom first
   unnest(test) %>%
   filter(p.value < 0.05)
   #mutate(p.adjust = p.adjust(p.value, method="fdr")) %>% #https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6099145/
@@ -194,7 +194,7 @@ composite_V4_sep_species <- composite_V4_sep_genera(composite_V4_sep, Species)
 sig_family_V4 <-
   composite_V4_sep_family %>%
   nest(data = -Family) %>%
-  mutate(test = map(.x= data, ~wilcox.test(total_genera_sum~ pot_soil, data=.x, exact = FALSE) %>% tidy)) %>% #run broom first
+  mutate(test = map(.x= data, ~wilcox.test(total_genera_sum~ potex_con, data=.x, exact = FALSE) %>% tidy)) %>% #run broom first
   unnest(test) %>%
   filter(p.value < 0.05)
 
@@ -253,24 +253,47 @@ composite_V4_pot_control <-
   labs(x= "Relative abundance (%)",y= NULL) +
   theme(axis.text.y = element_markdown())
 
-composite_V4_pot <-
+sig_family_all <- rbind(sig_family_V4, sig_family_V4_pot)
+sig_family_name <- distinct(sig_family_all, Family)
+
+composite_V4_combine_pot_con <-
   composite_V4_sep_family %>%
-  inner_join(sig_family_V4_pot, by="Family") %>%
+  inner_join(sig_family_name, by = "Family") %>%
   mutate(Family = str_remove(Family, "f__"),
          Family = str_replace(Family, "(.*)", "*\\1*")) %>%
-  filter(!is.na(pot)) %>%
   mutate(total_genera_sum = 100 * (total_genera_sum + 1/15000), # make it percentage and add a very small number
[…]

### 2024-02-11 004-taxa-relative-ab.R
reduce border lines in barplot

@@ -120,11 +120,13 @@ V4_bac_phyla_rel_ab <- # region number can be replaced
                           "unclassified", paste0(Domain,"; ",Phylum))) %>%
   mutate(Taxonomy = case_when(Taxonomy == "k__Bacteria; others" ~ "others",
                               Taxonomy == "k__Archaea; others" ~ "others", TRUE ~ as.character(Taxonomy))) %>%
-  ggplot(aes(x = label, y = count, fill = Taxonomy)) +
-  geom_bar(stat="identity", position = position_fill(reverse = TRUE), width = 0.8) + #"stack" shows original count; # fill
+  ggplot(aes(x = label, y = count, fill = Taxonomy, color = Taxonomy)) +
+  geom_bar(stat="identity", position = position_fill(reverse = TRUE), #"stack" shows original count; # fill
+           width = 0.8, linewidth = 0.1) + # linewidth to reduce the border lines
   scale_y_continuous(labels = scales::percent, expand = c(0.01, 0.01)) +
   scale_fill_manual(values = rev(phylum_colors_rel_mod)) + #breaks = V4_Phylum_group_order
-  guides(fill = guide_legend(title = "Taxonomy (phylum level)")) +
+  scale_color_manual(values = rev(phylum_colors_rel_mod)) +
+  guides(fill = guide_legend(title = "Taxonomy (phylum level)"), color = "none") +
   labs(x = NULL, y = "Relative Abundance") +
   theme(panel.background = element_blank(),
         panel.border = element_blank()) +

### 2024-02-12 003-rarefy-taxa-test.R
combine relative abundance plots

@@ -272,28 +272,14 @@ composite_V4_combine_pot_con <-
   scale_y_discrete(limits=rev) +
   labs(x= "Relative abundance (%)", y= NULL) +
   theme_minimal() +
-  theme(axis.text.y = element_markdown())
+  theme(axis.text.y = element_markdown(size= 12),
+        legend.title=element_text(size=12),
+        legend.text=element_text(size=12))
 
-library(patchwork)
 library(cowplot)
-library(png)
+V4_rel_sig_taxa_combine <-
+  plot_grid(V4_bac_phyla_rel_ab , composite_V4_combine_pot_con, # run 004-taxa-relative-ab to get the object
+            labels = "AUTO", ncol = 1)
 
-# use ggdraw from cowplot to plot the image
-plt1 <- readPNG(here::here("analysis", "figures", "03_rel_ab_V4.png")) %>%
-  cowplot::draw_image()
-
-# get dummy plot
-plt2 <- ggplot2::ggplot() +
-  geom_blank() +
-  theme(aspect.ratio = 4/3)
-
-# combine with patchwork
-plot_combined <- (plt2 + plt1 )/plt1  +
-  plot_layout(heights = c(1, 1)) +
-  plot_annotation(tag_levels = 'A')
-
-# save to png
-ggsave(filename = "plot_combined.png",
-       plot = plot_combined,
-       width = 4,
-       height = 4)
+ggsave(here::here("analysis","figures","V4_rel_sig_taxa_combine.png"),
+       width = 9.5, height = 7, dpi = 360, units = "in")

### 2024-02-12 004-taxa-relative-ab.R
adjust the text size in the plot

@@ -129,7 +129,9 @@ V4_bac_phyla_rel_ab <- # region number can be replaced
   guides(fill = guide_legend(title = "Taxonomy (phylum level)"), color = "none") +
   labs(x = NULL, y = "Relative Abundance") +
   theme(panel.background = element_blank(),
-        panel.border = element_blank()) +
+        panel.border = element_blank(),
+        legend.title=element_text(size=11),
+        legend.text=element_text(size=10)) +
   facet_grid(~group, scale= "free", space = "free")
 
 # export the df before plotting

### 2024-02-13 003-rarefy-taxa-test.R
work on significant test

@@ -79,7 +79,7 @@ composite_V6 <- composite_region(shared_V6_rarefy, taxonomy_V6, metadata_V6)
 composite_ITS1 <- composite_region(shared_ITS1_rarefy, taxonomy_ITS1, metadata_ITS1)
 
 # get significant genera
-sig_genera_V4 <-
+sig_genera_V4 <- # replace the region
   composite_V4 %>%
   nest(data = -taxonomy) %>% #c(-taxonomy, -otu)
   mutate(test = map(.x= data, ~wilcox.test(rel_abund~ pot, data=.x, exact = FALSE) %>% tidy)) %>% #run broom first
@@ -94,7 +94,7 @@ sig_genera_V1 <-
   nest(data = -taxonomy) %>% #c(-taxonomy, -otu)
   mutate(test = map(.x= data, ~wilcox.test(rel_abund~  pot_con, data=.x, exact = FALSE) %>% tidy)) %>% #run broom first
   unnest(test) %>%
-  filter(p.value < 0.1)
+  filter(p.value < 0.05)
 
 sig_genera_V6 <-
   composite_V6 %>%
@@ -108,7 +108,7 @@ sig_genera_ITS1 <-
   nest(data = -taxonomy) %>% #c(-taxonomy, -otu)
   mutate(test = map(.x= data, ~wilcox.test(rel_abund~  pot_con, data=.x, exact = FALSE) %>% tidy)) %>% #run broom first
   unnest(test) %>%
-  filter(p.value < 0.1)
+  filter(p.value < 0.05)
 
 # make a plot
 library(ggtext)
@@ -175,9 +175,12 @@ tax_clean_sep <- function(df){
 }
 
 composite_V4_sep <- tax_clean_sep(composite_V4)
+composite_V6_sep <- tax_clean_sep(composite_V6)
+composite_V1_sep <- tax_clean_sep(composite_V1)
+composite_ITS1_sep <- tax_clean_sep(composite_ITS1)
 
 # function for adding up relative abundance based on different level
-composite_V4_sep_genera <- function(df, col){ # col= Family, Genus etc.
+composite_sep_genera <- function(df, col){ # col= Family, Genus etc.
   df %>%
     group_by(Group, {{col}}) %>%
     mutate(total_genera_sum = sum(rel_abund)) %>%
@@ -186,47 +189,56 @@ composite_V4_sep_genera <- function(df, col){ # col= Family, Genus etc.
              potex_con, potin_con, potin_soil, potex_soil, pot_soil)
 }
 
[…]

### 2024-02-13 004-taxa-relative-ab.R
export data

@@ -24,6 +24,8 @@ V4_context <- region_context(context, "V4")
 V1_context <- region_context(context, "V1")
 V6_context <- region_context(context, "V6")
 ITS1_context <- region_context(context, "ITS1")
+ITS1_context <- ITS1_context %>%
+  unite("label", c(label, area), na.rm = TRUE, remove = FALSE)
 
 # rarefied taxonomy data with sample name and ASV name
 tax_rarefy_df <- function(df1, df2, df3){
@@ -135,8 +137,8 @@ V4_bac_phyla_rel_ab <- # region number can be replaced
   facet_grid(~group, scale= "free", space = "free")
 
 # export the df before plotting
-write.csv(V6_bac_phyla_rel_ab,
-          here::here("analysis", "data", "derived_data", "V6_bac_phyla_rel_ab.csv"),
+write.csv(V4_bac_phyla_rel_ab,
+          here::here("analysis", "data", "derived_data", "V4_bac_phyla_rel_ab.csv"),
                      row.names = FALSE)
 
 library(Cairo)
@@ -202,6 +204,11 @@ ITS1_bac_phyla_rel_ab <-
   guides(fill = guide_legend(title = "Taxonomy (class level)")) +
   labs(x = NULL, y = "Relative Abundance") +
   facet_grid(~type, scale= "free", space = "free")
+
+# export the df before plotting
+write.csv(ITS1_bac_phyla_rel_ab,
+          here::here("analysis", "data", "derived_data", "ITS1_bac_phyla_rel_ab.csv"),
+          row.names = FALSE)
 
 CairoPNG(here::here("analysis","figures","03_rel_ab_ITS1.png"), width = 9.5, height = 4, dpi = 360, units = "in")
 ITS1_bac_phyla_rel_ab

### 2024-02-13 analysis/paper/paper.qmd
work on results

--- 2024-02-09 analysis/paper/paper.qmd
@@
 
-In this study, our objective is to investigate the feasibility of using microbial communities as indicators of past food processing in pottery by focusing on the potsherds from the Guijiabao site in southwest China.
-As a long-term occupation site from Neolithic period to historical period, Guijiabao revealed abundant archaeological assemblages and remaines to explore the adaptation of farming communities in a mountain setting.
-The post-excavation analysis of Guijiabao is still in progress.
-This study serves as a pilot study to examine the pottery use from a perspective of microbial archaeology.
+This study serves as a pilot study to investigate the feasibility of using microbial communities as indicators of past food processing in pottery by focusing on the potsherds from the Guijiabao site in southwest China.
+As a long-term occupation site from Neolithic period to historical period, Guijiabao revealed abundant archaeological assemblages and remains to explore the adaptation of farming communities in a mountain setting.
 We aim to explore the differences in overall microbial communities, including bacterial and fungal communities, through a comparative analysis of different parts of ceramic samples and soil samples.
@@
 We hypothesized that the microbial composition in absorbed residues within the pottery matrix will exhibit distinctions relevant to food residues when compared to those on the surface of pottery and in surrounding soils.
+We also compared the resolution of V1, V4, and V6 16S rRNA regions to estimate microbial community diversity.
 While microorganisms offer indirect evidence of past food practices, our focus is on their potential to provide valuable information that complements existing scientific approaches to analyzing food residues.
@@
 V6_taxa_ab <- read.csv(here::here("analysis","data","derived_data","V6_bac_phyla_rel_ab.csv"))
+ITS1_taxa_ab <- read.csv(here::here("analysis","data","derived_data","ITS1_bac_phyla_rel_ab.csv"))
+
 V4_taxa_all_sample <- 
@@
 library(corrr)
-rel_ab_wider <- 
-  V4_taxa_all_sample %>% 
+# function to create a wide dataframe
+rel_ab_wider <- function(df){
+  df %>% 
   select(label, Phylum, each_phylum_sum) %>% #each_phylum_sum, rel_ab
@@
   select(-Phylum) 
+}
+
+rel_ab_wider_v4 <- rel_ab_wider(V4_taxa_all_sample)
 
 # compute correlation test
-cor_rel_ab <- rel_ab_wider %>%  correlate()
+cor_rel_ab_v4 <- rel_ab_wider_v4 %>%  correlate()
 
@@
 }
-p_value_cor <- colpair_map(rel_ab_wider, calc_cor_p_value)
+p_value_cor_v4 <- colpair_map(rel_ab_wider_v4, calc_cor_p_value)
 
@@
 }
-p_value_wil<- colpair_map(rel_ab_wider, calc_wil_p_value)
-
-# GLM
-calc_glm_p_value <- function(vec_a, vec_b){
-  x <- glm(vec_a ~ vec_b, family = poisson)
-  broom::tidy(x)$p.value[2]
-}
-p_value_glm <- colpair_map(rel_ab_wider, calc_glm_p_value)
+p_value_wil_v4 <- colpair_map(rel_ab_wider_v4, calc_wil_p_value)
 
@@
 V4_Firmi <- major_taxa(V4_taxa_ab, "p__Firmicutes")
-
-Bacilli <- V4_Firmi %>% filter(Class == "c__Bacilli")
-Clost <- V4_Firmi %>% filter(Class == "c__Clostridia")
-Nega <- V4_Firmi %>% filter(Class == "c__Negativicutes")
-
 V4_Actin <- major_taxa(V4_taxa_ab, "p__Actinobacteria")
@@
 
+V6_taxa_all_sample <- 
+  V6_taxa_ab %>%  
+  group_by(label) %>% 
+  mutate(total_phylum_sum = sum(count)) %>% 
+  ungroup() %>% 
+  group_by(Phylum, label) %>% 
+  mutate(each_phylum_sum = sum(count)) %>% 
+  ungroup() %>% 
+  distinct(label, Phylum, total_phylum_sum, each_phylum_sum) %>% 
+  mutate(rel_ab = round(100*each_phylum_sum/total_phylum_sum, 1))
+
+rel_ab_wider_v6 <- rel_ab_wider(V6_taxa_all_sample)
+p_value_wil_v6 <- colpair_map(rel_ab_wider_v6, calc_wil_p_value)
+pottery_v6 <-V6_taxa_all_sample %>% filter(str_detect(label, "pot"))
+Pates <- pottery_v6 %>% filter(Phylum == "p__Patescibacteria")
+V6_Pates <- major_taxa(V6_taxa_ab, "p__Patescibacteria")
+
+ITS1_taxa_all_sample <- 
+  ITS1_taxa_ab %>%  
+  group_by(label) %>% 
+  mutate(total_phylum_sum = sum(count)) %>% 
+  ungroup() %>% 
+  group_by(Phylum, label) %>% 
+  mutate(each_phylum_sum = sum(count)) %>% 
+  ungroup() %>% 
+  distinct(label, Phylum, total_phylum_sum, each_phylum_sum) %>% 
+  mutate(rel_ab = round(100*each_phylum_sum/total_phylum_sum, 1))
+
+rel_ab_wider_its1 <- rel_ab_wider(ITS1_taxa_all_sample)
+p_value_wil_its1 <- colpair_map(rel_ab_wider_its1, calc_wil_p_value)
+Ascom <- ITS1_taxa_all_sample %>% filter(Phylum == "p__Ascomycota")
+
 ```
@@
 ## Taxonomic composition 
-For the V4 region, bacterial composition at the phylum level in pottery interior samples was similar to those of pottery exterior samples and control samples, but slightly different from that of soil samples (@fig-V4-rel-ab).
+For the V4 region, bacterial composition at the phylum level in pottery interior samples was similar to those of pottery exterior samples and control samples, but slightly different from that of soil samples (@fig-V4-rel-ab: A).
 For Guijiabao pottery samples, *Firmicutes* (`r min(Firmi$rel_ab)`- `r max(Firmi$rel_ab)`%) was the most abundant phylum, followed by *Actinobacteria* (`r min(Actin$rel_ab)`- `r max(Actin$rel_ab)`%), *Deinococcus* (`r min(Deino$rel_ab)`- `r max(Deino$rel_ab)`%), and *Proteobacteria* (`r min(Prote$rel_ab)`- `r max(Prote$rel_ab)`%).
@@
 The soil samples from Guijiabao is predominated by *Firmicutes* (`r min(Firmi_soil$rel_ab)`- `r max(Firmi_soil$rel_ab)`%) and *Thaumarchaeota* (`r min(Thau_soil$rel_ab)`- `r max(Thau_soil$rel_ab)`%) belonging to the archaeal domain, while soil samples from Gaoshan is predominated by *Actinobacteria* (`r Actin_soil4$rel_ab`- `r Actin_soil5$rel_ab`%) with less abundant *Thaumarchaeota* (`r Thau_soil4$rel_ab`- `r Thau_soil5$rel_ab`%).
-The dominant classes include *Bacilli*, *Clostridia* and *Negativicutes* among *Firmicutes*; *Actinobacteria*, *Rubrobacteria*, *Thermoleophilia* among *Actinobacteria*, *Deinococci* in *Deinococcus*, and *Gammaproteobacteria*, *Alphaproteobacteria*, and *Deltaproteobacteria* among *Proteobacteria*.
 
+The dominant classes include *Bacilli*, *Clostridia*, and *Negativicutes* within *Firmicutes*; *Actinobacteria*, *Rubrobacteria*, *Thermoleophilia* within *Actinobacteria*, *Deinococci* in *Deinococcus*, and *Gammaproteobacteria*, *Alphaproteobacteria*, and *Deltaproteobacteria* within *Proteobacteria*.
 At the family level, significant differences were observed between pottery exterior samples and control samples, and between pottery exterior and interior samples.
 *Paenibacillaceae*, *Erysipelotrichaceae*, and *Xiphinematobacteraceae* were more abundant in pottery exterior samples than those in control samples (P <0.05).
-In addition to *Paenibacillaceae* and *Xiphinematobacteraceae*, *Phycisphaeraceae* was more abundant in pottery exterior samples than in pottery interior samples (P <0.05).
+In addition to *Paenibacillaceae* and *Xiphinematobacteraceae*, *Phycisphaeraceae* was also more abundant in pottery exterior samples than in pottery interior samples (P <0.05).
+At the genus level, *Candidatus_Xiphinematobacter* was more abundant in pottery exterior samples compared to pottery interior samples and control samples (P <0.01).
+Among the bacterial taxa showing significant differences, only *Paenibacillaceae* displayed relative abundances exceeding 10% on average, while others were below 1% (@fig-V4-rel-ab: B).
+
+The V6 region has similar bacterial communities at the phylum level as the V4 region across samples, with the exception of *Patescibacteria* (`r min(Pates$rel_ab)`- `r max(Pates$rel_ab)`%) in pottery samples, which was not detected in the reads of the V4 region.
+The dominant classes within *Patescibacteria* include *Parcubacteria*, *Microgenomatia*, and *Saccharimonadia*.
+We found a significant difference in the relative abundance of phyla between the pottery interior sample and soil samples from Gaoshan, but none for Guijiabao samples.
+At the family level, *Paenibacillaceae* was more abundant in pottery exterior samples than in pottery interior samples (P <0.05), while *Sphingomonas* was more abundant in pottery interior samples than in control samples.
+The bacterial communities at the phylum level for the V1 region are similar to those for the V4 and V6 regions, although there were no significant variations in composition between samples.
+For the ITS1 region of 18S rRNA, *Ascomycota* (`r min(Firmi$rel_ab)`- `r max(Firmi$rel_ab)`%) and *Basidiomycota* (`r min(Actin$rel_ab)`- `r max(Actin$rel_ab)`%) were the dominant phyla.
 
@@
 #| cache: true
-#| fig-cap: "The taxonomic composition of bacterial communities for the V4 region of pottery, soil, and control samples at the genus level."
+#| fig-cap: "A: The taxonomic composition of bacterial communities for the V4 region of pottery, soil, and control samples at the genus level.
+B: Bacterial communities at the family level differed significantly in relative abundance between pottery exterior and control samples, and between pottery exterior and interior samples."
 library(here)
-knitr::include_graphics(here("analysis", "figures", "03_rel_ab_V4.png"))
+knitr::include_graphics(here("analysis", "figures", "V4_rel_sig_taxa_combine.png"))
 ```

### 2024-02-14 analysis/paper/paper.qmd
work on relative abundance

@@
 p_value_wil_its1 <- colpair_map(rel_ab_wider_its1, calc_wil_p_value)
-Ascom <- ITS1_taxa_all_sample %>% filter(Phylum == "p__Ascomycota")
+Ascom_GJB <- ITS1_taxa_all_sample %>% filter(str_detect(label, "GJB") & Phylum == "p__Ascomycota")
+Ascom_GS <- ITS1_taxa_all_sample %>% filter(str_detect(label, "GS") & Phylum == "p__Ascomycota")
+Basid_GJB <- ITS1_taxa_all_sample %>% filter(str_detect(label, "GJB") & Phylum == "p__Basidiomycota")
+Basid_GS <- ITS1_taxa_all_sample %>% filter(str_detect(label, "GS") & Phylum == "p__Basidiomycota")
 
@@
 At the family level, *Paenibacillaceae* was more abundant in pottery exterior samples than in pottery interior samples (P <0.05), while *Sphingomonas* was more abundant in pottery interior samples than in control samples.
+V1 and ITS1 regions have higher proportions of unidentified reads than V4 and V6 regions.
 The bacterial communities at the phylum level for the V1 region are similar to those for the V4 and V6 regions, although there were no significant variations in composition between samples.
-For the ITS1 region of 18S rRNA, *Ascomycota* (`r min(Firmi$rel_ab)`- `r max(Firmi$rel_ab)`%) and *Basidiomycota* (`r min(Actin$rel_ab)`- `r max(Actin$rel_ab)`%) were the dominant phyla.
-
+For the ITS1 region of 18S rRNA, fungal communities at the phylum level were predominated by *Ascomycota* (`r min(Ascom_GJB$rel_ab)`- `r max(Ascom_GJB$rel_ab)`% for Guijiabao samples and `r min(Ascom_GS$rel_ab)`- `r max(Ascom_GS$rel_ab)`% for Gaoshan samples) and *Basidiomycota* (`r min(Basid_GJB$rel_ab)`- `r max(Basid_GJB$rel_ab)`% for Guijiabao samples and `r min(Basid_GS$rel_ab)`- `r max(Basid_GS$rel_ab)`% for Gaoshan samples.
 ```{r}

### 2024-03-01 002-rarefy-beta.R
work on Jaccard measures

@@ -35,12 +35,13 @@ context_tidy <-
          pot_soil = case_when(type== "pot" ~ "pot", type== "soil" ~ "soil"),
          pot_con = case_when(type== "pot" ~ "pot", type== "control" ~ "control"),
          potex_con = case_when(group == "pot-exterior" ~ "pot-exterior", type== "control" ~ "control"),
-         potin_con = case_when(group == "pot-interior" ~ "pot-interior", type== "control" ~ "control"))
+         potin_con = case_when(group == "pot-interior" ~ "pot-interior", type== "control" ~ "control")) %>%
+  mutate(order = parse_number(label))
 
-V4_meta<- filter(context_tidy, Region == "V4")
-V1_meta<- filter(context_tidy, Region == "V1")
-V6_meta<- filter(context_tidy, Region == "V6")
-ITS1_meta<- filter(context_tidy, Region == "ITS1")
+V4_meta<- filter(context_tidy, Region == "V4") %>% arrange(`adpater no`)
+V1_meta<- filter(context_tidy, Region == "V1") %>% arrange(`adpater no`)
+V6_meta<- filter(context_tidy, Region == "V6") %>% arrange(`adpater no`)
+ITS1_meta<- filter(context_tidy, Region == "ITS1") %>% arrange(`adpater no`)
 
 # wider dataframe based on rarefied OTUs (use this to make all the order consistent)
 rarefy_clean_wider <- function(df){
@@ -104,10 +105,16 @@ V4_stress_b <- stress(rarefy_V4_clean_wider, "bray")
 V1_stress_b <- stress(rarefy_V1_clean_wider, "bray")
 V6_stress_b <- stress(rarefy_V6_clean_wider, "bray")
 ITS1_stress_b <- stress(rarefy_ITS1_clean_wider, "bray")
-V4_pv_bray <- c(V4_pv_bray, V4_stress_b)
-V1_pv_bray <- c(V1_pv_bray, V1_stress_b)
-V6_pv_bray <- c(V6_pv_bray, V6_stress_b)
-ITS1_pv_bray <- c(ITS1_pv_bray, ITS1_stress_b)
+
+V4_F_bray <- format(round(V4_bray_perm$F[1], 3))
+V1_F_bray <- format(round(V1_bray_perm$F[1], 3))
[…]

### 2024-03-02 analysis/paper/paper.qmd
work on beta diversity

@@
 For the ITS1 region of 18S rRNA, fungal communities at the phylum level were predominated by *Ascomycota* (`r min(Ascom_GJB$rel_ab)`- `r max(Ascom_GJB$rel_ab)`% for Guijiabao samples and `r min(Ascom_GS$rel_ab)`- `r max(Ascom_GS$rel_ab)`% for Gaoshan samples) and *Basidiomycota* (`r min(Basid_GJB$rel_ab)`- `r max(Basid_GJB$rel_ab)`% for Guijiabao samples and `r min(Basid_GS$rel_ab)`- `r max(Basid_GS$rel_ab)`% for Gaoshan samples.
+
 ```{r}
@@
 ## Beta diversity analysis
+We used principal coordinates analysis (PCoA) based on binary Jaccard and Bray-Curtis dissimilarity to evaluate any structural differences between bacterial and fungal communities in four groups: control, pottery interior, pottery exterior, and soil samples.
+The bacterial communities identified by the V1, V4, and V6 region sequences indicated a significant difference between four sample groups for their presence and abundance according to PERMANOVA tests (P < 0.01), while there were no significant differences in the fungal communities.
+A further pairwise PERMANOVA tests demonstrated bacterial communities in pottery interior and exterior samples differed significantly from soil samples (P < 0.05) based on V4 and V6 region sequences, but not on V1 region sequences.
+Although there were no significant differences in beta diversity between pottery interior, pottery exterior, and control samples, the Jaccard PCoA plot (V4 region) showed a higher intra-variability between pottery samples from Guijiabao and Gaoshan, as well as soil samples from two sites according to the Jaccard PCoA plot (V6 region) and the Bray-Curtis PCoA plots (V4 and V6 regions).
 
-Alpha diversity 
-Beta diversity
-Relative abundance 
 Venn diagrams

### 2024-03-03 002-rarefy-beta.R
add shape

@@ -36,6 +36,7 @@ context_tidy <-
          pot_con = case_when(type== "pot" ~ "pot", type== "control" ~ "control"),
          potex_con = case_when(group == "pot-exterior" ~ "pot-exterior", type== "control" ~ "control"),
          potin_con = case_when(group == "pot-interior" ~ "pot-interior", type== "control" ~ "control")) %>%
+  mutate(site = case_when(site == "GJB" ~ "Guijiabao", site == "CDG" ~ "Gaoshan", TRUE ~ site)) %>%
   mutate(order = parse_number(label))
 
 V4_meta<- filter(context_tidy, Region == "V4") %>% arrange(`adpater no`)
@@ -144,9 +145,9 @@ ITS1_bray_eig <- round(100*ITS1_bray$eig/sum(ITS1_bray$eig), 2)
 # create a PCoA bray plot
 library(ggrepel)
 library(ggtext)
-pcoa_bray_plot <-
+V4_pcoa_bray_plot <-
   V4_pcoa_bray_for_plot %>%
-  ggplot(aes(x= axis_1, y= axis_2, color= group)) +
+  ggplot(aes(x= axis_1, y= axis_2, color= group, shape = site)) +
   geom_point(size= 3) +
   geom_text_repel(aes(label = order), size = 3.5,  hjust = 2,
                   min.segment.length = 0.2, show.legend = FALSE) +
@@ -161,7 +162,22 @@ pcoa_bray_plot <-
         legend.text=element_text(size=rel(1.2)),
         strip.text.x = element_text(size=4),strip.text.y = element_text(size=4))
 
-ggsave(here::here("analysis", "figures", "PcoA_bray_rare.png"), width = 6.5, height = 4.5, units = "in")
+V6_pcoa_bray_plot <-
+  V6_pcoa_bray_for_plot %>%
+  ggplot(aes(x= axis_1, y= axis_2, color= group, shape = site)) +
+  geom_point(size= 3) +
+  geom_text_repel(aes(label = order), size = 3.5,  hjust = 2,
+                  min.segment.length = 0.2, show.legend = FALSE) +
+  annotate("text", -Inf, Inf, label = paste(c("italic(R)^2==" , "P<", "F=="), V6_pv_bray),
+           hjust = c(-0.2, -0.2, -0.2), vjust = c(1.25, 3, 4.25), parse = T ) +
+  scale_colour_viridis_d(option = "magma", begin= 0.1, end= 0.8, direction = -1) + #change color by option = "magma"
[…]

### 2024-03-03 analysis/paper/paper.qmd
add PCoA plots

--- 2024-02-14 analysis/paper/paper.qmd
@@
 ## Beta diversity analysis
+We used principal coordinates analysis (PCoA) based on binary Jaccard and Bray-Curtis dissimilarity to evaluate any structural differences between bacterial and fungal communities in four groups: control, pottery interior, pottery exterior, and soil samples.
+The bacterial communities identified by the V1, V4, and V6 region sequences showed a significant difference between four sample groups for their presence and abundance according to PERMANOVA tests (P < 0.01), while there were no significant differences in the fungal communities (@fig-V4-V6-PCoA).
+A further pairwise PERMANOVA tests showed bacterial communities in pottery interior and exterior samples differed significantly from soil samples (P < 0.05) based on the V4 and V6 region sequences, but not on the V1 region sequences.
+Although there were no significant differences in beta diversity between pottery interior, pottery exterior, and control samples, the Jaccard PCoA plot (the V4 region) indicated a higher intra-variability between pottery samples from Guijiabao and Gaoshan, as well as soil samples from two sites according to the Jaccard PCoA plot (the V6 region) and the Bray-Curtis PCoA plots (the V4 and V6 regions).
 
-Alpha diversity 
-Beta diversity
-Relative abundance 
+```{r}
+#| label: fig-V4-V6-PCoA
+#| echo: false
+#| cache: true
+#| fig-cap: "A: PCoA plots of bacterial communities based on the sequences of V4 region.
+B: PCoA plots of bacterial communities based on the sequences of V6 region.
+The percentage variability explained by each principal component is indicated on the axes"
+knitr::include_graphics(here("analysis", "figures", "V4_V6_combine_PcoA.png"))
+```
+
 Venn diagrams

### 2024-03-04 004-taxa-relative-ab.R
edit plots and combine

@@ -24,8 +24,8 @@ V4_context <- region_context(context, "V4")
 V1_context <- region_context(context, "V1")
 V6_context <- region_context(context, "V6")
 ITS1_context <- region_context(context, "ITS1")
-ITS1_context <- ITS1_context %>%
-  unite("label", c(label, area), na.rm = TRUE, remove = FALSE)
+#ITS1_context <- ITS1_context %>%
+  #unite("label", c(label, area), na.rm = TRUE, remove = FALSE)
 
 # rarefied taxonomy data with sample name and ASV name
 tax_rarefy_df <- function(df1, df2, df3){
@@ -122,7 +122,7 @@ V4_bac_phyla_rel_ab <- # region number can be replaced
                           "unclassified", paste0(Domain,"; ",Phylum))) %>%
   mutate(Taxonomy = case_when(Taxonomy == "k__Bacteria; others" ~ "others",
                               Taxonomy == "k__Archaea; others" ~ "others", TRUE ~ as.character(Taxonomy))) %>%
-  ggplot(aes(x = label, y = count, fill = Taxonomy, color = Taxonomy)) +
+  ggplot(aes(x = label, y = count, fill = Taxonomy, color = Taxonomy)) + #set color to make border invisible
   geom_bar(stat="identity", position = position_fill(reverse = TRUE), #"stack" shows original count; # fill
            width = 0.8, linewidth = 0.1) + # linewidth to reduce the border lines
   scale_y_continuous(labels = scales::percent, expand = c(0.01, 0.01)) +
@@ -136,15 +136,12 @@ V4_bac_phyla_rel_ab <- # region number can be replaced
         legend.text=element_text(size=10)) +
   facet_grid(~group, scale= "free", space = "free")
 
+ggsave(here::here("analysis","figures","03_rel_ab_V4.png"), width = 10, height = 3.5, dpi = 360, units = "in")
+
 # export the df before plotting
 write.csv(V4_bac_phyla_rel_ab,
           here::here("analysis", "data", "derived_data", "V4_bac_phyla_rel_ab.csv"),
                      row.names = FALSE)
-
-library(Cairo)
-CairoPNG(here::here("analysis","figures","03_rel_ab_V4.png"), width = 10, height = 4, dpi = 360, units = "in")
-V4_bac_phyla_rel_ab
-dev.off()
 
[…]

### 2024-03-06 003-rarefy-taxa-test.R
code edit

@@ -101,7 +101,7 @@ sig_genera_V6 <-
   nest(data = -taxonomy) %>% #c(-taxonomy, -otu)
   mutate(test = map(.x= data, ~wilcox.test(rel_abund~  pot_con, data=.x, exact = FALSE) %>% tidy)) %>% #run broom first
   unnest(test) %>%
-  filter(p.value < 0.1)
+  filter(p.value < 0.05)
 
 sig_genera_ITS1 <-
   composite_ITS1 %>%
@@ -203,18 +203,18 @@ composite_ITS1_sep_genus <- composite_sep_genera(composite_ITS1_sep, Genus)
 composite_ITS1_sep_species <- composite_sep_genera(composite_ITS1_sep, Species)
 
 # get significant genera at family level between pottery and control
-sig_family_ITS1 <-
-  composite_ITS1_sep_family %>% # replace the region
+sig_family_V4 <-
+  composite_V4_sep_family %>% # replace the region
   nest(data = -Family) %>%
   mutate(test = map(.x= data, ~wilcox.test(total_genera_sum~ potex_con, data=.x, exact = FALSE) %>% tidy)) %>% #run broom first
   unnest(test) %>%
   filter(p.value < 0.05)
 
 # get significant genera at family level between pottery
-sig_family_ITS1_pot <-
-  composite_ITS1_sep_family %>%
+sig_family_V6_pot <-
+  composite_V6_sep_family %>%
   nest(data = -Family) %>%
-  mutate(test = map(.x= data, ~wilcox.test(total_genera_sum~ pot, data=.x, exact = FALSE) %>% tidy)) %>% #run broom first
+  mutate(test = map(.x= data, ~wilcox.test(total_genera_sum~ potin_con, data=.x, exact = FALSE) %>% tidy)) %>% #run broom first
   unnest(test) %>%
   filter(p.value < 0.05)
 
@@ -287,6 +287,9 @@ composite_V4_combine_pot_con <-
   theme(axis.text.y = element_markdown(size= 12),
         legend.title=element_text(size=12),
         legend.text=element_text(size=12))
+
+ggsave(here::here("analysis","figures","V4_rel_sig_taxa_combine.png"),
+       width = 9.5, height = 7, dpi = 360, units = "in")
 
 library(cowplot)
 V4_rel_sig_taxa_combine <-

### 2024-03-07 003-rarefy-taxa-test.R
work on significant tests

@@ -203,17 +203,24 @@ composite_ITS1_sep_genus <- composite_sep_genera(composite_ITS1_sep, Genus)
 composite_ITS1_sep_species <- composite_sep_genera(composite_ITS1_sep, Species)
 
 # get significant genera at family level between pottery and control
-sig_family_V4 <-
+sig_family_V4_pot_ex <-
   composite_V4_sep_family %>% # replace the region
   nest(data = -Family) %>%
   mutate(test = map(.x= data, ~wilcox.test(total_genera_sum~ potex_con, data=.x, exact = FALSE) %>% tidy)) %>% #run broom first
   unnest(test) %>%
   filter(p.value < 0.05)
 
+sig_family_V4_pot <-
+  composite_V4_sep_family %>% # replace the region
+  nest(data = -Family) %>%
+  mutate(test = map(.x= data, ~wilcox.test(total_genera_sum~ pot, data=.x, exact = FALSE) %>% tidy)) %>% #run broom first
+  unnest(test) %>%
+  filter(p.value < 0.05)
+
 # get significant genera at family level between pottery
 sig_family_V6_pot <-
-  composite_V6_sep_family %>%
-  nest(data = -Family) %>%
+  composite_V6_sep_genus %>%
+  nest(data = -Genus) %>%
   mutate(test = map(.x= data, ~wilcox.test(total_genera_sum~ potin_con, data=.x, exact = FALSE) %>% tidy)) %>% #run broom first
   unnest(test) %>%
   filter(p.value < 0.05)

### 2024-03-08 analysis/paper/paper.qmd
add a code chunk

@@
 Basid_GS <- ITS1_taxa_all_sample %>% filter(str_detect(label, "GS") & Phylum == "p__Basidiomycota")
-
 ```
@@
 ## Taxonomic composition 
-For the V4 region, bacterial composition at the phylum level in pottery interior samples was similar to those of pottery exterior samples and control samples, but slightly different from that of soil samples (@fig-V4-rel-ab: A).
+For the V4 region, bacterial composition at the phylum level in pottery interior samples was similar to those of pottery exterior samples and control samples, but slightly different from that of soil samples (@fig-rel-ab: A).
 For Guijiabao pottery samples, *Firmicutes* (`r min(Firmi$rel_ab)`- `r max(Firmi$rel_ab)`%) was the most abundant phylum, followed by *Actinobacteria* (`r min(Actin$rel_ab)`- `r max(Actin$rel_ab)`%), *Deinococcus* (`r min(Deino$rel_ab)`- `r max(Deino$rel_ab)`%), and *Proteobacteria* (`r min(Prote$rel_ab)`- `r max(Prote$rel_ab)`%).
@@
 
-The dominant classes include *Bacilli*, *Clostridia*, and *Negativicutes* within *Firmicutes*; *Actinobacteria*, *Rubrobacteria*, *Thermoleophilia* within *Actinobacteria*, *Deinococci* in *Deinococcus*, and *Gammaproteobacteria*, *Alphaproteobacteria*, and *Deltaproteobacteria* within *Proteobacteria*.
+```{r abundance-significant}
+#| label: abundance-significant
+#| echo: false
+#| cache: true
+library(broom)
+library(Hmisc)
+# function to read data
+shared_rarefy_long <- function(filename, vec){
+  read.csv(here::here("analysis","data","derived_data", filename), sep = vec) %>%
+    select(Group, starts_with("ASV")) %>%
+    pivot_longer(-Group, names_to="otu", values_to="count")
+}
+
+shared_V4_rarefy <-shared_rarefy_long("V4_shared_format.1.subsample.60216.csv", ",")
+shared_V1_rarefy <-shared_rarefy_long("V1_shared_format.1.subsample.13658.csv", "")
+shared_V6_rarefy <-shared_rarefy_long("V6_shared_format.1.subsample.28402.csv", ",")
+shared_ITS1_rarefy <-shared_rarefy_long("ITS1_shared_format.1.subsample.387.csv", ",")
+
+# function to tidy taxonomy data
+tidy_taxonomy <- function(filename){
+  read.table(here::here("analysis","data","raw_data",filename), header = T, sep="\t") %>%
+  select(OTU.ID, taxonomy) %>%
+  rename(otu = OTU.ID) %>%
+  mutate(taxonomy = str_replace_all(taxonomy, "\\(\\d+\\)", ""),
+         taxonomy = str_replace(taxonomy, "; c__; o__; f__; g__; s__", ""),
+         taxonomy = str_replace(taxonomy, "; o__; f__; g__; s__", ""),
+         taxonomy = str_replace(taxonomy, "; f__; g__; s__", ""),
+         taxonomy = str_replace(taxonomy, "; g__; s__", ""),
+         taxonomy = str_replace(taxonomy, "(; s__)([^(; s__)]*)$", ""),
+         taxonomy = str_replace(taxonomy, "(; f__)([^(; f__)]*)$", ""),
+         taxonomy = str_replace(taxonomy, "(; g__)([^(; g__)]*)$", ""))
+  }
+
+taxonomy_V4 <- tidy_taxonomy("V4_asv_table.txt")
+taxonomy_V1 <- tidy_taxonomy("V1_asv_table.txt")
+taxonomy_V6 <- tidy_taxonomy("V6_asv_table.txt")
+taxonomy_ITS1 <- tidy_taxonomy("ITS1_asv_table.txt")
+
+# function to read metadata for each region
+meta_region_extract <- function(filename, vec){
+  read_excel(here::here("analysis","data", "raw_data", filename)) %>%
+    filter(Region == vec) %>%
+    rename(Group = `adpater no`) %>%
+    mutate(type= str_extract(Group, "[[:alpha:]]+(?=\\d)"), type = tolower(type)) %>%
+    mutate(group = if_else(!is.na(area)&!area%in% c("belly", "outside belly"),
+                           paste0(type, "-", area), type),
+           pot = case_when(group == "pot-inside" ~ "interior", group == "pot-outside" ~"exterior"),
+           potex_soil = case_when(group == "pot-outside" ~ "pot-exterior", type== "soil" ~ "soil"),
+           potin_soil = case_when(group == "pot-inside" ~ "pot-interior", type== "soil" ~ "soil"),
+           pot_soil = case_when(type== "pot" ~ "pot", type== "soil" ~ "soil"),
+           pot_con = case_when(type== "pot" ~ "pot", type== "con" ~ "control"),
+           potex_con = case_when(group == "pot-outside" ~ "pot-exterior", type== "con" ~ "control"),
+           potin_con = case_when(group == "pot-inside" ~ "pot-interior", type== "con" ~ "control")) %>%
+    relocate(Group, type, group, sample, area, site)
+}
+
+metadata_V4 <- meta_region_extract("GJB_amp_meta.xlsx", "V4")
+metadata_V1 <- meta_region_extract("GJB_amp_meta.xlsx", "V1")
+metadata_V6 <- meta_region_extract("GJB_amp_meta.xlsx", "V6")
+metadata_ITS1 <- meta_region_extract("GJB_amp_meta.xlsx", "ITS1")
+
+# function to calculate relative abundance for taxa level, not ASV level
+composite_region <- function(df1, df2, df3){
+  inner_join(df1, df2, by= "otu") %>%
+    group_by(Group, taxonomy) %>%
+    dplyr::summarize(count = sum(count), .groups="drop") %>%
+    group_by(Group) %>%
+    mutate(rel_abund = count / sum(count)) %>%
+    ungroup() %>%
+    select(-count) %>%
+    inner_join(., df3, by = "Group") %>%
+    filter(!taxonomy == "k__Bacteria")
+}
+
+composite_V4 <- composite_region(shared_V4_rarefy, taxonomy_V4, metadata_V4)
+composite_V1 <- composite_region(shared_V1_rarefy, taxonomy_V1, metadata_V1)
+composite_V6 <- composite_region(shared_V6_rarefy, taxonomy_V6, metadata_V6)
+composite_ITS1 <- composite_region(shared_ITS1_rarefy, taxonomy_ITS1, metadata_ITS1)
+
+# function to separate the taxonomy data, also in 000-tidy-data
+tax_clean_sep <- function(df){
+  df %>%
+    separate(taxonomy, into = c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), sep=";") %>%
+    mutate(Phylum = str_trim(Phylum), Class = str_trim(Class), 
+           Order = str_trim(Order),
+           Family = str_trim(Family), Genus = str_trim(Genus), 
+           Species = str_trim(Species)) %>%
+    mutate(Domain = ifelse(Domain == "Unassigned", "unclassified", Domain),
+           Phylum = ifelse(is.na(Phylum), "unclassified", Phylum),
+           Class = ifelse(is.na(Class), "unclassified", Class),
+           Order = ifelse(is.na(Order), "unclassified", Order),
+           Family = ifelse(is.na(Family), "unclassified", Family),
+           Genus = ifelse(is.na(Genus), "unclassified", Genus),
+           Species = ifelse(is.na(Species), "unclassified", Species)) %>%
+    mutate(Phylum = ifelse(Phylum == "p__WPS_2", "p__Eremiobacterota", Phylum),
+           Phylum = ifelse(Phylum == "p__Deinococcus_Thermus","p__Deinococcus", Phylum))
+}
+
+composite_V4_sep <- tax_clean_sep(composite_V4)
+composite_V6_sep <- tax_clean_sep(composite_V6)
+composite_V1_sep <- tax_clean_sep(composite_V1)
+composite_ITS1_sep <- tax_clean_sep(composite_ITS1)
+
+# function for adding up relative abundance based on different level
+composite_sep_genera <- function(df, col){ # col= Family, Genus etc.
+  df %>%
+    group_by(Group, {{col}}) %>%
+    mutate(total_genera_sum = sum(rel_abund)) %>%
+    ungroup() %>%
+    distinct(Group, {{col}}, group, total_genera_sum, pot_con, pot,
+             potex_con, potin_con, potin_soil, potex_soil, pot_soil)
+}
+
+composite_V4_sep_family <- composite_sep_genera(composite_V4_sep, Family)
+composite_V4_sep_genus <- composite_sep_genera(composite_V4_sep, Genus)
+composite_V4_sep_species <- composite_sep_genera(composite_V4_sep, Species)
+composite_V6_sep_family <- composite_sep_genera(composite_V6_sep, Family)
+composite_V6_sep_genus <- composite_sep_genera(composite_V6_sep, Genus)
+composite_V6_sep_species <- composite_sep_genera(composite_V6_sep, Species)
+composite_V1_sep_family <- composite_sep_genera(composite_V1_sep, Family)
+composite_V1_sep_genus <- composite_sep_genera(composite_V1_sep, Genus)
+composite_V1_sep_species <- composite_sep_genera(composite_V1_sep, Species)
+composite_ITS1_sep_family <- composite_sep_genera(composite_ITS1_sep, Family)
+composite_ITS1_sep_genus <- composite_sep_genera(composite_ITS1_sep, Genus)
+composite_ITS1_sep_species <- composite_sep_genera(composite_ITS1_sep, Species)
+```
+
+```{r plots-abundance-significant}
+#| label: plots-abundance-significant
+#| echo: false
+#| cache: true
+# get significant genera at family level between pottery and control
+sig_family_V4_potex_con <-
+  composite_V4_sep_family %>% # replace the region
+  nest(data = -Family) %>%
+  mutate(test = map(.x= data, ~wilcox.test(total_genera_sum~ potex_con, data=.x, exact = FALSE) %>% tidy)) %>% #run broom first
+  unnest(test) %>%
+  filter(p.value < 0.05)
+
+sig_family_V4_pot <-
+  composite_V4_sep_family %>% # replace the region
+  nest(data = -Family) %>%
+  mutate(test = map(.x= data, ~wilcox.test(total_genera_sum~ pot, data=.x, exact = FALSE) %>% tidy)) %>% #run broom first
+  unnest(test) %>%
+  filter(p.value < 0.05)
+
+sig_genus_V4_pot <-
+  composite_V4_sep_genus %>% # replace the region
+  nest(data = -Genus) %>%
+  mutate(test = map(.x= data, ~wilcox.test(total_genera_sum~ pot, data=.x, exact = FALSE) %>% tidy)) %>% #run broom first
+  unnest(test) %>%
+  filter(p.value < 0.05)
+
+V4_sig_family_all <- rbind(sig_family_V4_potex_con, sig_family_V4_pot)
+V4_sig_family_name <- distinct(V4_sig_family_all, Family)
+
+composite_V4_combine_pot_con <-
+  composite_V4_sep_family %>%
+  inner_join(V4_sig_family_name, by = "Family") %>%
+  mutate(Family = str_remove(Family, "f__"),
+         Family = str_replace(Family, "(.*)", "*\\1*")) %>%
+  mutate(total_genera_sum = 100 * (total_genera_sum + 1/15000), # make it percentage and add a very small number
+         pot = factor(group, levels = c("con", "pot-inside", "pot-outside", "soil"))) %>%
+  ggplot(aes(x= total_genera_sum, y = Family, fill= group)) +
+  geom_boxplot(aes(fill = group), color = "gray60", show.legend = TRUE) +
+  scale_x_log10() +
+  scale_fill_viridis_d(breaks = c("con", "pot-inside", "pot-outside", "soil"),
+                       labels = c("control","pot-interior", "pot-exterior", "soil"),
+                       direction = -1) +
+  scale_y_discrete(limits=rev) +
+  labs(x= "Relative abundance (%)", y= NULL) +
+  theme_minimal() +
+  theme(axis.text.y = element_markdown(size= 12),
+        legend.title=element_text(size=12),
+        legend.text=element_text(size=12))
+
+ggsave(here::here("analysis","figures","V4_rel_sig_taxa_combine.png"),
+       width = 9.5, height = 7, dpi = 360, units = "in")
+
+sig_family_V6_pot <-
+  composite_V6_sep_family %>% # replace the region
+  nest(data = -Family) %>%
+  mutate(test = map(.x= data, ~wilcox.test(total_genera_sum~ pot, data=.x, exact = FALSE) %>% tidy)) %>% #run broom first
+  unnest(test) %>%
+  filter(p.value < 0.05)
+
+sig_genus_V6_potin_con <-
+  composite_V6_sep_genus %>% # replace the region
+  nest(data = -Genus) %>%
+  mutate(test = map(.x= data, ~wilcox.test(total_genera_sum~ potin_con, data=.x, exact = FALSE) %>% tidy)) %>% #run broom first
+  unnest(test) %>%
+  filter(p.value < 0.05)
+
+V6_sig_family_all <- rbind(sig_family_V6_pot, sig_family_V6_potin_con, sig_family_V6_potex_con)
+V6_sig_family_name <- distinct(V6_sig_family_all, Family)
+V6_sig_genus_name <- sig_genus_V6_potin_con$Genus
+
+composite_V6_combine_pot_con <-
+  composite_V6_sep_genus %>%
+  inner_join(sig_genus_V6_potin_con, by = "Genus") %>%
+  mutate(Genus = str_remove(Genus, "g__"),
+         Genus = str_replace(Genus, "(.*)", "*\\1*")) %>%
+  mutate(total_genera_sum = 100 * (total_genera_sum + 1/15000), # make it percentage and add a very small number
+         pot = factor(group, levels = c("con", "pot-inside", "pot-outside", "soil"))) %>%
+  ggplot(aes(x= total_genera_sum, y = Genus, fill= group)) +
+  geom_boxplot(aes(fill = group), color = "gray60", show.legend = TRUE) +
+  scale_x_log10() +
+  scale_fill_viridis_d(breaks = c("con", "pot-inside", "pot-outside", "soil"),
+                       labels = c("control","pot-interior", "pot-exterior", "soil"),
+                       direction = -1) +
+  scale_y_discrete(limits=rev) +
+  labs(x= "Relative abundance (%)", y= NULL) +
+  theme_minimal() +
+  theme(axis.text.y = element_markdown(size= 12),
+        legend.title=element_text(size=12),
+        legend.text=element_text(size=12))
+
+```
+
+The dominant classes include *Bacilli*, *Clostridia*, and *Negativicutes* within *Firmicutes*; *Actinobacteria*, *Rubrobacteria*, *Thermoleophilia* within *Actinobacteria*, *Deinococci* within *Deinococcus*, and *Gammaproteobacteria*, *Alphaproteobacteria*, and *Deltaproteobacteria* within *Proteobacteria*.
 At the family level, significant differences were observed between pottery exterior samples and control samples, and between pottery exterior and interior samples.
@@
 At the genus level, *Candidatus_Xiphinematobacter* was more abundant in pottery exterior samples compared to pottery interior samples and control samples (P <0.01).
-Among the bacterial taxa showing significant differences, only *Paenibacillaceae* displayed relative abundances exceeding 10% on average, while others were below 1% (@fig-V4-rel-ab: B).
+Among the bacterial taxa showing significant differences, only *Paenibacillaceae* in the pottery exterior sample displayed relative abundances exceeding 10% on average, while others were below 1% (figure).
 
-The V6 region has similar bacterial communities at the phylum level as the V4 region across samples, with the exception of *Patescibacteria* (`r min(Pates$rel_ab)`- `r max(Pates$rel_ab)`%) in pottery samples, which was not detected in the reads of the V4 region.
+The V6 region has similar bacterial communities at the phylum level as the V4 region, with the exception of *Patescibacteria* (`r min(Pates$rel_ab)`- `r max(Pates$rel_ab)`%) in pottery samples, which was not detected in the reads of the V4 region (@fig-rel-ab: B).
 The dominant classes within *Patescibacteria* include *Parcubacteria*, *Microgenomatia*, and *Saccharimonadia*.
-We found a significant difference in the relative abundance of phyla between the pottery interior sample and soil samples from Gaoshan, but none for Guijiabao samples.
-At the family level, *Paenibacillaceae* was more abundant in pottery exterior samples than in pottery interior samples (P <0.05), while *Sphingomonas* was more abundant in pottery interior samples than in control samples.
-V1 and ITS1 regions have higher proportions of unidentified reads than V4 and V6 regions.
-The bacterial communities at the phylum level for the V1 region are similar to those for the V4 and V6 regions, although there were no significant variations in composition between samples.
+We found a significant difference in the relative abundance of phyla between the pottery interior sample and soil samples from Gaoshan (P < 0.05), but none for Guijiabao samples.
+At the family level, *Paenibacillaceae* and *67_14* was more abundant in pottery exterior samples than in pottery interior samples (P <0.05).
+At the genus level, *Sphingomonas* was more abundant in the pottery interior samples than in the control samples (P <0.05).
+
+The V1 and ITS1 regions have higher proportions of unidentified reads than V4 and V6 regions (@fig-rel-ab: C-D).
+The bacterial communities at the phylum level for the V1 region are similar to those for the V4 and V6 regions, but there were no significant variations in composition between samples.
 For the ITS1 region of 18S rRNA, fungal communities at the phylum level were predominated by *Ascomycota* (`r min(Ascom_GJB$rel_ab)`- `r max(Ascom_GJB$rel_ab)`% for Guijiabao samples and `r min(Ascom_GS$rel_ab)`- `r max(Ascom_GS$rel_ab)`% for Gaoshan samples) and *Basidiomycota* (`r min(Basid_GJB$rel_ab)`- `r max(Basid_GJB$rel_ab)`% for Guijiabao samples and `r min(Basid_GS$rel_ab)`- `r max(Basid_GS$rel_ab)`% for Gaoshan samples.
 
+B: Bacterial communities at the family level differed significantly in relative abundance between pottery exterior and control samples, and between pottery exterior and interior samples
+
 ```{r}
-#| label: fig-V4-rel-ab
+#| label: fig-rel-ab
 #| echo: false
 #| cache: true
-#| fig-cap: "A: The taxonomic composition of bacterial communities for the V4 region of pottery, soil, and control samples at the genus level.
-B: Bacterial communities at the family level differed significantly in relative abundance between pottery exterior and control samples, and between pottery exterior and interior samples."
+#| fig-cap: "A, B, C: The taxonomic composition of bacterial communities in pottery interior, pottery exterior, soil, and control samples based on sequences of the V4, V6, and V1 regions.
+D: The taxonomic composition of bacterial communities based on sequences of the ITS1 region."
 library(here)
-knitr::include_graphics(here("analysis", "figures", "V4_rel_sig_taxa_combine.png"))
+knitr::include_graphics(here("analysis", "figures", "all_rel_ab_combine.png"))
 ```

### 2024-03-10 analysis/paper/paper.qmd
edit code chunk and add a plot

@@
   unnest(test) %>%
+  filter(p.value < 0.05) %>% 
+  filter(!Genus == "unclassified")
+
+V4_sig_family_all <- rbind(sig_family_V4_potex_con, sig_family_V4_pot) %>% slice(1,2,5)
+
+# join the original data of significance test 
+composite_V4_genus_for_plot <-
+  composite_V4_sep_genus %>%
+  inner_join(sig_genus_V4_pot, by = "Genus") %>%
+  mutate(Genus = str_remove(Genus, "g__")) %>% 
+  rename(Taxon = Genus)
+
+composite_V4_family_for_plot <-
+  composite_V4_sep_family %>%
+  inner_join(V4_sig_family_all, by = "Family") %>%
+  mutate(Family = str_remove(Family, "f__")) %>% 
+  rename(Taxon = Family)
+
+# V4 plot
+library(ggtext)
+composite_V4_family_genus_com_plot <-
+  rbind(composite_V4_genus_for_plot, composite_V4_family_for_plot) %>%
+  mutate(Taxon = str_replace(Taxon, "(.*)", "*\\1*"),
+         Taxon = str_replace(Taxon, "\\*Candidatus_(.*)\\*", "Candidatus<br>*\\1*")) %>%
+  mutate(total_genera_sum = 100 * (total_genera_sum + 1/15000), # make it percentage and add a very small number
+         pot = factor(group, levels = c("con", "pot-inside", "pot-outside", "soil"))) %>%
+  ggplot(aes(x= total_genera_sum, y = Taxon, fill= group)) +
+  geom_boxplot(aes(fill = group), color = "gray60", show.legend = TRUE) +
+  scale_x_log10() +
+  scale_fill_viridis_d(breaks = c("con", "pot-inside", "pot-outside", "soil"),
+                       labels = c("control","pot-interior", "pot-exterior", "soil"), direction = -1) +
+  scale_y_discrete(limits=rev) +
+  labs(x= "Relative abundance (%)", y= NULL) +
+  theme_minimal() +
+  theme(axis.text.y = element_markdown(size= 12), legend.position = "none")
+
+sig_family_V6_pot <-
+  composite_V6_sep_family %>% # replace the region
+  nest(data = -Family) %>%
+  mutate(test = map(.x= data, ~wilcox.test(total_genera_sum~ pot, data=.x, exact = FALSE) %>% tidy)) %>% #run broom first
+  unnest(test) %>%
   filter(p.value < 0.05)
 
-V4_sig_family_all <- rbind(sig_family_V4_potex_con, sig_family_V4_pot)
-V4_sig_family_name <- distinct(V4_sig_family_all, Family)
+sig_family_V6_potin_con <-
+  composite_V6_sep_family %>% # replace the region
+  nest(data = -Family) %>%
+  mutate(test = map(.x= data, ~wilcox.test(total_genera_sum~ potin_con, data=.x, exact = FALSE) %>% tidy)) %>% #run broom first
+  unnest(test) %>%
+  filter(p.value < 0.05)
 
-composite_V4_combine_pot_con <-
-  composite_V4_sep_family %>%
-  inner_join(V4_sig_family_name, by = "Family") %>%
-  mutate(Family = str_remove(Family, "f__"),
-         Family = str_replace(Family, "(.*)", "*\\1*")) %>%
+sig_genus_V6_potin_con <-
+  composite_V6_sep_genus %>% # replace the region
+  nest(data = -Genus) %>%
+  mutate(test = map(.x= data, ~wilcox.test(total_genera_sum~ potin_con, data=.x, exact = FALSE) %>% tidy)) %>% #run broom first
+  unnest(test) %>%
+  filter(p.value < 0.05)
+
+V6_sig_family_all <- rbind(sig_family_V6_pot, sig_family_V6_potin_con) %>% slice(1:2)
+
+# join the original data of significance test 
+composite_V6_genus_for_plot <-
+  composite_V6_sep_genus %>%
+  inner_join(sig_genus_V6_potin_con, by = "Genus") %>%
+  mutate(Genus = str_remove(Genus, "g__")) %>% 
+  rename(Taxon = Genus)
+
+composite_V6_family_for_plot <-
+  composite_V6_sep_family %>%
+  inner_join(V6_sig_family_all, by = "Family") %>%
+  mutate(Family = str_remove(Family, "f__")) %>% 
+  rename(Taxon = Family)
+
+# V6 plot
+composite_V6_family_genus_com_plot <-
+  rbind(composite_V6_genus_for_plot, composite_V6_family_for_plot) %>%
+  mutate(Taxon = str_replace(Taxon, "(.*)", "*\\1*")) %>%
   mutate(total_genera_sum = 100 * (total_genera_sum + 1/15000), # make it percentage and add a very small number
          pot = factor(group, levels = c("con", "pot-inside", "pot-outside", "soil"))) %>%
-  ggplot(aes(x= total_genera_sum, y = Family, fill= group)) +
+  ggplot(aes(x= total_genera_sum, y = Taxon, fill= group)) +
   geom_boxplot(aes(fill = group), color = "gray60", show.legend = TRUE) +
@@
 
-ggsave(here::here("analysis","figures","V4_rel_sig_taxa_combine.png"),
-       width = 9.5, height = 7, dpi = 360, units = "in")
+library(cowplot)
+V4_V6_rel_sig_taxa_combine <-
+  plot_grid(composite_V4_family_genus_com_plot, composite_V6_family_genus_com_plot, rel_widths = c(0.8, 1), labels = "AUTO", ncol = 2)
 
-sig_family_V6_pot <-
-  composite_V6_sep_family %>% # replace the region
-  nest(data = -Family) %>%
-  mutate(test = map(.x= data, ~wilcox.test(total_genera_sum~ pot, data=.x, exact = FALSE) %>% tidy)) %>% #run broom first
-  unnest(test) %>%
-  filter(p.value < 0.05)
-
-sig_genus_V6_potin_con <-
-  composite_V6_sep_genus %>% # replace the region
-  nest(data = -Genus) %>%
-  mutate(test = map(.x= data, ~wilcox.test(total_genera_sum~ potin_con, data=.x, exact = FALSE) %>% tidy)) %>% #run broom first
-  unnest(test) %>%
-  filter(p.value < 0.05)
-
-V6_sig_family_all <- rbind(sig_family_V6_pot, sig_family_V6_potin_con, sig_family_V6_potex_con)
-V6_sig_family_name <- distinct(V6_sig_family_all, Family)
-V6_sig_genus_name <- sig_genus_V6_potin_con$Genus
-
-composite_V6_combine_pot_con <-
-  composite_V6_sep_genus %>%
-  inner_join(sig_genus_V6_potin_con, by = "Genus") %>%
-  mutate(Genus = str_remove(Genus, "g__"),
-         Genus = str_replace(Genus, "(.*)", "*\\1*")) %>%
-  mutate(total_genera_sum = 100 * (total_genera_sum + 1/15000), # make it percentage and add a very small number
-         pot = factor(group, levels = c("con", "pot-inside", "pot-outside", "soil"))) %>%
-  ggplot(aes(x= total_genera_sum, y = Genus, fill= group)) +
-  geom_boxplot(aes(fill = group), color = "gray60", show.legend = TRUE) +
-  scale_x_log10() +
-  scale_fill_viridis_d(breaks = c("con", "pot-inside", "pot-outside", "soil"),
-                       labels = c("control","pot-interior", "pot-exterior", "soil"),
-                       direction = -1) +
-  scale_y_discrete(limits=rev) +
-  labs(x= "Relative abundance (%)", y= NULL) +
-  theme_minimal() +
-  theme(axis.text.y = element_markdown(size= 12),
-        legend.title=element_text(size=12),
-        legend.text=element_text(size=12))
-
+ggsave(here::here("analysis","figures","V4_V6_rel_sig_taxa_combine.png"),
+       width = 10, height = 4, dpi = 360, units = "in")
 ```
@@
 At the family level, significant differences were observed between pottery exterior samples and control samples, and between pottery exterior and interior samples.
-*Paenibacillaceae*, *Erysipelotrichaceae*, and *Xiphinematobacteraceae* were more abundant in pottery exterior samples than those in control samples (P <0.05).
+*Paenibacillaceae*, *Erysipelotrichaceae*, and *Xiphinematobacteraceae* were more abundant in pottery exterior samples than those in control samples (P <0.05; @fig-rel-ab-significant-plot: A).
 In addition to *Paenibacillaceae* and *Xiphinematobacteraceae*, *Phycisphaeraceae* was also more abundant in pottery exterior samples than in pottery interior samples (P <0.05).
-At the genus level, *Candidatus_Xiphinematobacter* was more abundant in pottery exterior samples compared to pottery interior samples and control samples (P <0.01).
-Among the bacterial taxa showing significant differences, only *Paenibacillaceae* in the pottery exterior sample displayed relative abundances exceeding 10% on average, while others were below 1% (figure).
+At the genus level, Candidatus *Xiphinematobacter* was more abundant in pottery exterior samples compared to pottery interior samples and control samples (P <0.01).
+Among the bacterial taxa showing significant differences, only *Paenibacillaceae* in the pottery exterior sample displayed relative abundances exceeding 10% on average, while others were below 1%.
 
@@
 We found a significant difference in the relative abundance of phyla between the pottery interior sample and soil samples from Gaoshan (P < 0.05), but none for Guijiabao samples.
-At the family level, *Paenibacillaceae* and *67_14* was more abundant in pottery exterior samples than in pottery interior samples (P <0.05).
+At the family level, *Paenibacillaceae* and *67_14* was more abundant in pottery exterior samples than in pottery interior samples (P <0.05; @fig-rel-ab-significant-plot: B).
 At the genus level, *Sphingomonas* was more abundant in the pottery interior samples than in the control samples (P <0.05).
@@
 
-B: Bacterial communities at the family level differed significantly in relative abundance between pottery exterior and control samples, and between pottery exterior and interior samples
-
 ```{r}
@@
 #| echo: false
-#| cache: true
 #| fig-cap: "A, B, C: The taxonomic composition of bacterial communities in pottery interior, pottery exterior, soil, and control samples based on sequences of the V4, V6, and V1 regions.
@@
 knitr::include_graphics(here("analysis", "figures", "all_rel_ab_combine.png"))
+```
+
+```{r}
+#| label: fig-rel-ab-significant-plot
+#| echo: false
+#| fig-cap: "A: Bacterial communities at the family and genus level differed significantly in relative abundance between pottery exterior and control samples, and between pottery exterior and interior samples based on sequences of the V4 region.
+B: Bacterial communities at the family and genus level differed significantly in relative abundance between pottery exterior and interior samples, and between pottery interior and control samples based on sequences of the V6 region."
+library(here)
+knitr::include_graphics(here("analysis", "figures", "V4_V6_rel_sig_taxa_combine.png"))
 ```

### 2024-03-12 005-venn_diagram.R
edit venn diagrams

@@ -45,23 +45,53 @@ ITS1_pot_ex_list<- as.vector(ITS1_pot_ex_df$Species)
 ITS1_soil_list <- as.vector(ITS1_soil_df$Species)
 
 # Venn diagram
-V4_venn <- list(control= V4_con_list , "pot-interior"= V4_pot_in_list,
+V4_venn <- list(control= V4_con_list, "pot-interior"= V4_pot_in_list,
                 "pot-exterior"= V4_pot_ex_list, soil= V4_soil_list)
 
-V1_venn <- list(control= V1_con_list , "pot-interior"= V1_pot_in_list,
+V1_venn <- list(control= V1_con_list, "pot-interior"= V1_pot_in_list,
                 "pot-exterior"= V1_pot_ex_list, soil= V1_soil_list)
 
-V6_venn <- list(control= V6_con_list , "pot-interior"= V6_pot_in_list,
+V6_venn <- list(control= V6_con_list, "pot-interior"= V6_pot_in_list,
                 "pot-exterior"= V6_pot_ex_list, soil= V6_soil_list)
 
-ITS1_venn <- list(control= ITS1_con_list , "pot-interior"= ITS1_pot_in_list,
+ITS1_venn <- list(control= ITS1_con_list, "pot-interior"= ITS1_pot_in_list,
                 "pot-exterior"= ITS1_pot_ex_list, soil= ITS1_soil_list)
 
-ggvenn(ITS1_venn, fill_color = c("#0073C2FF", "#EFC000FF", "#33A02C", "#CD534CFF"),
-       stroke_size = 0.5, set_name_size = 4)+
+ggplot(d) +
+  geom_venn(aes(A = `Set 1`, B = `Set 2`)) +
+  coord_fixed() +
+  theme_void()
+
+V4_venn_diagram <-
+  ggvenn(V4_venn, fill_color = c("#0073C2FF", "#EFC000FF", "#33A02C", "#CD534CFF"),
+       stroke_size = 0.5, set_name_size = 4, fill_alpha = 0.3)+
   theme(plot.background = element_rect(fill = "white", colour = "white"),
         strip.text = element_text(size= 12))
-ggsave(here::here("analysis", "figures", "ITS1_venndiag.png"), width = 5, height = 4, units = "in")
+
+V6_venn_diagram <-
+  ggvenn(V6_venn, fill_color = c("#0073C2FF", "#EFC000FF", "#33A02C", "#CD534CFF"),
[…]

### 2024-03-23 005-venn_diagram.R
save venn diagram based on ASV

@@ -139,6 +139,12 @@ V1_taxa_venn_diagram <- venn_diagram(V1_taxa_venn)
 V6_taxa_venn_diagram <- venn_diagram(V6_taxa_venn)
 ITS1_taxa_venn_diagram <- venn_diagram(ITS1_taxa_venn)
 
+all_taxa_venn_diagram <-
+  plot_grid(V4_taxa_venn_diagram, V6_taxa_venn_diagram, V1_taxa_venn_diagram, ITS1_taxa_venn_diagram,
+            ncol = 2, labels = c('V4', 'V6', 'V1', 'ITS1'))
+
+ggsave(here::here("analysis", "figures", "all_taxa_venn_diagram.png"), width = 10, height = 7.5, units = "in")
+
 # check the unique ones for each group
 unique_ASV <- function(target_list, list1, list2, list3, df) {
   setdiff(target_list, c(list1, list2, list3)) %>%

### 2024-03-26 analysis/paper/paper.qmd
work on unique taxa

@@
 At the genus level, *Sphingomonas* was more abundant in the pottery interior samples than in the control samples (P <0.05).
-
 The V1 and ITS1 regions have higher proportions of unidentified reads than V4 and V6 regions (@fig-rel-ab: C-D).
@@
 
-Venn diagrams
-The abundant taxa at ASV level & the abundant taxa at genus level
+## Shared and unique microbial communities
+Pairwise comparison of taxonomy of each amplicon sequence variant (ASV) among different sample groups was conducted to investigate any unique microbial communities particularly in pottery samples.
+For the sequence data of 16S rRNA regions, V6 sequences show higher number of unique taxonomy (46, 10%) in pottery interior samples compared to those of V1 sequences (27, 19.3%) and V4 sequences (7, 1.1%), wheres the sequence of 18S rRNA based on ITS1 region presents no specific taxonomy in pottery interior samples (@fig-venn-diagram).
+Similar patterns were observed for pottery exterior samples, where V6 sequences present a higher number of unique taxonomy (42, 9.1%), followed by V1 sequences (11, 7.9%), V4 sequences (8, 1.3%), and ITS1 sequences (4, 9.8%).
+A close look at the unique taxa
+
+```{r}
+#| label: fig-venn-diagram
+#| echo: false
+#| cache: true
+#| fig-cap: "Venn diagrams based on reads from each regions analyzed in this study showing shared and specific species for bacterial and fungal communities for four sample groups"
+knitr::include_graphics(here("analysis", "figures", "all_taxa_venn_diagram.png"))
+```
 
